---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---
```{r echo=FALSE}

source("./src/custom.R")

```
 
 

# 🎯  Aim and content

::: {.callout-note icon=false}
## Overview

- General data visualization: *Good graphics*
- R-Projects for reproducible research (quick intro)
- Create basic scientific figures using ggplot2   
  - Applying simple modifications   
  - Figure legends and annotations 
  - Combining and arranging multiple figures 
  - Advanced boxplots
- Exporting high-quality figures (vector images) 
- QnA - Bring your problem

:::




# Schedule 

- Total: 7h 
- 09:00 - 12:30 (3.5h)
- Break (1h)
- 13:30 - 17:00 (3.5h)





```{r, echo=FALSE}
# 
# library("data.table")
# schedule <- data.table( 
# 					Timeslot= c(
# 						"09:00 -- 09:20", #"1 Intro", 
# 						"09:20 -- 09:50", #"2 First Plot",
# 						"09:50 -- 10:20", #"Exercise (1)",
# 						"10:20 -- 10:40", #"3 Plot types",
# 						"10:40 -- 11:10", #"Exercise (2) + short break",
# 						"11:10 -- 11:30", #"4 Layers "
# 						"11:30 -- 12:00", #"5 Colors, Shapes, Linetypes",
# 						"12:00 -- 12:30", #"Exercise (3)"
# 						"12:30 -- 13:30", #Lunch break
# 						"13:30 -- 14:00", #"6 Scientific figures",	
# 						"14:00 -- 14:20", #"Exercise (4)"
# 						"14:20 -- 14:50", #"7 Facets and patchwork",	
# 						"14:50 -- 15:20", #"Exercise (5) + short break"	
# 						"15:20 -- 15:30", #"8 Exporting vector graphics",	
# 						"15:30 -- 16:00", #"9 Optional topic",	
# 						"16:00 -- 17:00"  #"Q&A",		
# 						),
# 					                        
# 
# 					Topic = c(
# 						"1 Intro", 
# 						"2 First Plot",
# 						"Exercise (1)",
# 						"3 Plot types",
# 						"Exercise (2) + short break",
# 						"4 Layers ",
# 						"5 Colors, Shapes, Linetypes ",
# 						"Exercise (3)",
# 						"Lunch break",						
# 						"6 Scientific figures",	
# 						"Exercise (4)",
# 						"7 Facets and patchwork",	
# 						"Exercise (5) + short break",	
# 						"8 Exporting vector graphics + Exercise (6)",	
# 						"9 Optional topic",	
# 						"Q & A"
# 						)
# 					)
# 
# knitr::kable(schedule)

```



# Exercises / R-Environment

- Local R/R-Studio installation with R version 4.5.x or later   
- <https://rstudio.cloud/>  

## Packages

```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")

# install/load packages using pacman package
pacman::p_load("tidyverse")  	  # includes ggplot for plotting
pacman::p_load("rdwd")          # get DWD climate data (optional)
pacman::p_load("patchwork")     # align multiple plots
pacman::p_load("ggforce")       # zoom facet, annotation
pacman::p_load("ggthemes")      # (optional) themes for ggplot2
pacman::p_load("cowplot")      # (optional) themes for ggplot2
pacman::p_load("gghalves")      # half-box-plot etc.
pacman::p_load("ggrain")        # rain cloud plots
pacman::p_load("lvplot")        # boxenplot/letter value plot
pacman::p_load("ggridges")      # ridgeline plots
pacman::p_load("ggpubr")        # helper functions
pacman::p_load("ggdist")        # distribution plots
pacman::p_load("scico")         # scientific color maps
pacman::p_load("cols4all")      # color maps + evaluation 
pacman::p_load("directlabels")  # smart automatic direct labelling
pacman::p_load("geomtextpath")  # label along lines

```



# Additional resources

- <https://ggplot2-book.org/index.html>   
- <https://ggplot2.tidyverse.org/>  
- <https://r4ds.hadley.nz/data-visualize> 

#  Session Info

```{r}

sessionInfo()

```
