---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r format(Sys.time(), '%Y')`"
geometry: paperheight=20cm,paperwidth=30cm,margin=1cm
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: hide
  # html_document:
  #   #code_folding: hide
  #   toc: true
  #   toc_depth: 4
  #   theme: united
  #   highlight: zenburn
  #   number_sections: true
  #   toc_float: 
  #     collapsed: false
#      smooth_scroll: false
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: true
    highlight: zenburn
---

```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )

```
 
# Overview     

 
- Create basic scientific figures using ggplot2 
- Applying simple modifications 
- Figure legends and annotations 
- Combining and arranging multiple figures 
- Exporting high-quality figures (vector images) 
- Optional: Advanced boxplots
- Optional: Interactive plots in HTML documents 
- Optional: Data structure and management for plotting 



## Vote for optional topic


# Schedule 

- Total: 6h 
- 08:30 - 12:00 (3.5h)
- Break (1h)
- 13:00 - 15:30 (2.5h)






```{r}

library("data.table")
schedule <- data.table( 
					Timeslot= c(
						"08:30 -- 08:45", #"1 Intro", 
						"08:45 -- 09:00", #"2 First Plot",
						"09:00 -- 09:30", #"Exercise (1)",
						"09:30 -- 09:45", #"3 Plot types",
						"09:45 -- 10:15", #"Exercise (2) + short break",
						"10:15 -- 10:30", #"4 Layers "
						"10:30 -- 11:00", #"5 Colors, Shapes, Linetypes",
						"11:00 -- 12:00", #"Exercise (3)"
						"12:00 -- 13:00", #break
						"13:00 -- 13:30", #"6 Scientific figures",	
						"13:30 -- 13:45", #"Exercise (4)"
						"13:45 -- 14:00", #"7 Facets and patchwork",	
						"14:00 -- 14:30", #"Exercise (5) + short break"	
						"14:30 -- 14:45", #"8 Exporting vector graphics",	
						"14:45 -- 15:00", #"9 Optional topic",	
						"15:00 -- 15:30"  #"Q&A",		
						),
					                        

					Topic = c(
						"1 Intro", 
						"2 First Plot",
						"Exercise (1)",
						"3 Plot types",
						"Exercise (2) + short break",
						"4 Layers ",
						"5 Colors, Shapes, Linetypes",
						"Exercise (3)",
						"Break",
						"6 Scientific figures",	
						"Exercise (4)",
						"7 Facets and patchwork",	
						"Exercise (5) + short break",	
						"8 Exporting vector graphics + Exercise (6)",	
						"9 Optional topic",	
						"Q & A"
						)
					)

knitr::kable(schedule)




```



# Exercises / R-Environment

- Local R/R-Studio installation with R version 4.2.x or later
- https://rstudio.cloud/
- https://mybinder.org/v2/gh/dschmidt-hgu/ggplot2-course/main
- https://mybinder.org/v2/gh/dschmidt-hgu/ggplot2-course/main?urlpath=rstudio

## Packages

```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")

# install/load packages using pacman package
pacman::p_load("rmarkdown")     # to compile RMD files
pacman::p_load("rmdformats") 	  # to compile RMD files
pacman::p_load("ggplot2")  		  # plotting
pacman::p_load("data.table")    # data management
pacman::p_load("rdwd")          # get DWD climate data (optional)
pacman::p_load("patchwork")     # align multiple plots
pacman::p_load("scales")        # pretty scales
pacman::p_load("ggforce")       # zoom facet, annotation
pacman::p_load("sf")            # req. for gganmiate example
pacman::p_load("transformr")    # req. for gganimate example
pacman::p_load("gganimate")     # animate ggplot2s
pacman::p_load("ggthemes")      # themes for ggplot2
pacman::p_load("rgl")           # 3D plotting
pacman::p_load("plotly")        # Interactive graphs
pacman::p_load("gghalves")      # half-box-plot etc.
pacman::p_load("lvplot")        # boxenplot/letter value plot
pacman::p_load("ggridges")      # ridgeline plots
pacman::p_load("ggpubr")        # helper functions
pacman::p_load("ggdist")        # distribution plots


```



# Additional resources

- https://ggplot2-book.org/index.html
- https://ggplot2.tidyverse.org/


#  Session Info

```{r}

sessionInfo()

```
