---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
---
```{r echo=FALSE}

source("./src/custom.R")

```
 
# Overview     

 
- Create basic scientific figures using ggplot2 
- Applying simple modifications 
- Figure legends and annotations 
- Combining and arranging multiple figures 
- Exporting high-quality figures (vector images) 
- Optional: Advanced boxplots
- Optional: Interactive plots in HTML documents 



## Vote for optional topic

<br><br><br><br>


# Schedule 

- Total: 7h 
- 09:00 - 12:30 (3.5h)
- Break (1h)
- 13:30 - 17:00 (3.5h)






```{r, echo=FALSE}

library("data.table")
schedule <- data.table( 
					Timeslot= c(
						"09:00 -- 09:20", #"1 Intro", 
						"09:20 -- 09:50", #"2 First Plot",
						"09:50 -- 10:20", #"Exercise (1)",
						"10:20 -- 10:40", #"3 Plot types",
						"10:40 -- 11:10", #"Exercise (2) + short break",
						"11:10 -- 11:30", #"4 Layers "
						"11:30 -- 12:00", #"5 Colors, Shapes, Linetypes",
						"12:00 -- 12:30", #"Exercise (3)"
						"12:30 -- 13:30", #Lunch break
						"13:30 -- 14:00", #"6 Scientific figures",	
						"14:00 -- 14:20", #"Exercise (4)"
						"14:20 -- 14:50", #"7 Facets and patchwork",	
						"14:50 -- 15:20", #"Exercise (5) + short break"	
						"15:20 -- 15:30", #"8 Exporting vector graphics",	
						"15:30 -- 16:00", #"9 Optional topic",	
						"16:00 -- 17:00"  #"Q&A",		
						),
					                        

					Topic = c(
						"1 Intro", 
						"2 First Plot",
						"Exercise (1)",
						"3 Plot types",
						"Exercise (2) + short break",
						"4 Layers ",
						"5 Colors, Shapes, Linetypes ",
						"Exercise (3)",
						"Lunch break",						
						"6 Scientific figures",	
						"Exercise (4)",
						"7 Facets and patchwork",	
						"Exercise (5) + short break",	
						"8 Exporting vector graphics + Exercise (6)",	
						"9 Optional topic",	
						"Q & A"
						)
					)

knitr::kable(schedule)




```



# Exercises / R-Environment

- Local R/R-Studio installation with R version 4.3.x or later   
- <https://rstudio.cloud/>  
- <https://mybinder.org/v2/gh/dschmidt-hgu/ggplot2-course/main>   
- <https://mybinder.org/v2/gh/dschmidt-hgu/ggplot2-course/main?urlpath=rstudio>   

## Packages

```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")

# install/load packages using pacman package
pacman::p_load("ggplot2")  		  # plotting
pacman::p_load("data.table")    # data management
pacman::p_load("rdwd")          # get DWD climate data (optional)
pacman::p_load("patchwork")     # align multiple plots
pacman::p_load("scales")        # (pretty) scales + helpers
pacman::p_load("ggforce")       # zoom facet, annotation
#pacman::p_load("sf")            # req. for gganmiate example
pacman::p_load("transformr")    # req. for gganimate example
pacman::p_load("gganimate")     # animate ggplot2s
pacman::p_load("ggthemes")      # themes for ggplot2
pacman::p_load("rgl")           # 3D plotting
pacman::p_load("plotly")        # Interactive graphs
pacman::p_load("gghalves")      # half-box-plot etc.
pacman::p_load("lvplot")        # boxenplot/letter value plot
pacman::p_load("ggridges")      # ridgeline plots
pacman::p_load("ggpubr")        # helper functions
pacman::p_load("ggdist")        # distribution plots
pacman::p_load("skimr")         # complex data summary (optional)
pacman::p_load("hexbin")        # for geom_hex
pacman::p_load("scico")         # scientific color maps
pacman::p_load("cols4all")      # color maps + evaluation 
pacman::p_load("ggtext")        # more flexible text formatting in ggplot2
pacman::p_load("directlabels")  # smart automatic direct labelling
pacman::p_load("geomtextpath")  # label along lines

```



# Additional resources

- <https://ggplot2-book.org/index.html>
- <https://ggplot2.tidyverse.org/>


#  Session Info

```{r}

sessionInfo()

```
