---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}

source("./src/custom.R")

```


# 001: Data visualization

# 🎯 Learning objectives

::: {.callout-note icon=false}
## Why and how - Data visualization

**... understanding concepts:**

*     

**... tool usage:**

*   use your **data visualization literacy** to spot *bad graphics* 

:::




<br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Why visualize? - A: Exploratory data analysis

::: {.callout-tip icon=false}
## 👉 

**How would you visualize a scatterplot based on the following regression line and correlation coefficient?**

-  $y = 3 + 0.5x$ 
-  $r=0.82$	 

::: 

<br><br><br><br><br><br><br><br><br><br><br><br><br><br>




```{r echo=FALSE, message=FALSE}
pacman::p_load("tidyverse")
anscombe <- datasets::anscombe
anscombe.long <- data.frame(x=unlist(anscombe[,substr(colnames(anscombe),1,1)=="x"])
                           ,y=unlist(anscombe[,substr(colnames(anscombe),1,1)=="y"])
                           ,dataset=factor(rep(1:4,each=nrow(anscombe)))
                           )


ggplot(anscombe.long, aes(x=x,y=y))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,fullrange=TRUE)+
  facet_wrap(dataset~.,labeller="label_both")+
  theme_minimal()+
  ggtitle("Anscombe's quartet")+
  ggpubr::stat_cor(label.x = min(anscombe.long$x)
                  ,label.y = max(anscombe.long$x)
                  ,cor.coef.name="cor"
                  ,aes(label = after_stat(r.label))
                  ,size=2.75) +
  ggpubr::stat_regline_equation(label.x = min(anscombe.long$x)
                               ,label.y = max(anscombe.long$x)-2
                               ,size=2.75)+
  coord_fixed()




```

## *Anscombe's quartet* 

- Anscombe, F. J. (1973). "Graphs in Statistical Analysis". American Statistician. 27 (1): 17–21. doi: 10.1080/00031305.1973.10478966. JSTOR 2682899.

```{r, echo=FALSE}
anscombe.summary <- anscombe.long %>% 
                    group_by(dataset) %>%
                    summarize(
                      "mean_x"=mean(x)
                    ,"var_x"=var(x)
                    ,"mean_y"=round(mean(y),2)
                    ,"var_y"=round(var(y),3)
                    ,"corr(x,y)"=round(cor(x,y),3)
                    ,"lm"=paste0("y = ",sprintf(round(coef(lm(y~x)),2)[1],fmt="%1.2f")," + ",sprintf(round(coef(lm(y~x)),3)[2],fmt="%1.3f"),"x")
                    )

knitr::kable(anscombe.summary)


knitr::kable(anscombe)

```


## More illustrations on data patterns and correlation coefficients

- <https://janhove.github.io/teaching/2016/11/21/what-correlations-look-like>


::: color-box 
::: 

<br>
<br><br>

**Never trust summary statistics alone; always visualize your data.**

Alberto Cairo @ [twitter.com](https://twitter.com/albertocairo/status/770267777169035264)


<br><br><br>

**Graphics reveal data**

Tufte, Edward (1983). The Visual Display of Quantitative Information. Cheshire, Connecticut: Graphics Press. ISBN 0-9613921-4-2.



::: color-box 
::: 


# Why visualize? - B: Explanatory data visualization


A. **Exploratory Visualization (For YOU):**
   * **Goal:** Find patterns, spot outliers, check data quality.
   * **Look:** Can be messy, quick, and lacks perfect labels.

   
B. **Explanatory Visualization (For OTHERS):**
   * **Goal:** Communicate a specific finding to a specific audience (scientist, grower, public).
   * **Look:** Clean, clearly labeled, free of "chartjunk," with annotations guiding the eye.



::: color-box 
::: 



# "Excellence in statistical graphics ...

consists of complex ideas communicated with **clarity**, **precision**, and **efficiency**."


- **Clarity**
  - Lack of ambiguity and confusion   
- **Precision**   
  - Truthful results   
  - Distortion‐free presentation 
- **Efficiency**   
  - Minimal "chartjunk"   
  - Maximize "data-to-ink"-ratio   
  

<br>
Tufte, Edward (1983). The Visual Display of Quantitative Information. Cheshire, Connecticut: Graphics Press. ISBN 0-9613921-4-2.


::: color-box 
::: 



## Clarity?


[How to Lie with Statistics: Stand Your Ground and Gun Deaths](https://thesocietypages.org/socimages/2014/12/28/how-to-lie-with-statistics-stand-your-ground-and-gun-deaths/)

- ambiguity and confusion   


::: color-box 
::: 


## Precision?


[Social Distortion in Charts: The Venezuela Presidential Election 2013](https://datavizblog.com/2013/04/21/social-distortion-in-charts-the-venezuela-presidential-election-2013/)


::: color-box 
::: 



## Efficiency?


[Data Looks Better Naked - Remove to Improve: The Data-Ink Ratio Edition](https://darkhorsevisualization.com/blog/data-looks-better-naked)



::: color-box 
::: 




## *"Ugly, bad, and wrong figures"*


[Claus O. Wilke: Fundamentals of Data Visualization](https://clauswilke.com/dataviz/introduction.html)


::: color-box 
::: 

<br><br><br>


# ***Design is about efficient communication, not about making things pretty***

[2016 Markus Püschel, Computer Science, ETH Zürich](https://people.inf.ethz.ch/markusp/teaching/guides/guide-presentations-new.pdf)

<br><br>

::: color-box 
::: 

# Additional resources

- [Anscome's quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet)  
- [Claus O. Wilke: Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)  




#  Session Info

```{r}

sessionInfo()

```
