---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r format(Sys.time(), '%Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 002: First plot




# R and RStudio

## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
	
pacman::p_load("data.table")
pacman::p_load("ggplot2")


```




# Why visualizing data - Exploratory data analysis

```{r echo=FALSE, message=FALSE}

anscombe <- datasets::anscombe
anscombe.long <- data.table(x=unlist(anscombe[,substr(colnames(anscombe),1,1)=="x"])
                           ,y=unlist(anscombe[,substr(colnames(anscombe),1,1)=="y"])
                           ,dataset=factor(rep(1:4,each=nrow(anscombe)))
                           )


ggplot(anscombe.long, aes(x=x,y=y))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,fullrange=TRUE)+
  facet_wrap(dataset~.,labeller="label_both")+
  theme_minimal()+
  ggtitle("Anscombe's quartet")+
  ggpubr::stat_cor(label.x = min(anscombe.long$x)
                  ,label.y = max(anscombe.long$x)
                  ,cor.coef.name="cor"
                  ,aes(label = ..r.label..)
                  ,size=2.75) +
  ggpubr::stat_regline_equation(label.x = min(anscombe.long$x)
                               ,label.y = max(anscombe.long$x)-2
                               ,size=2.75)+
  coord_fixed()




```

## *Anscombe's quartet* 

- Anscombe, F. J. (1973). "Graphs in Statistical Analysis". American Statistician. 27 (1): 17–21. doi:10.1080/00031305.1973.10478966. JSTOR 2682899.

```{r, echo=FALSE}

knitr::kable(anscombe)

anscombe.summary <- anscombe.long[,list("mean_x"=mean(x)
                    ,"var_x"=var(x)
                    ,"mean_y"=round(mean(y),2)
                    ,"var_y"=round(var(y),3)
                    ,"corr(x,y)"=round(cor(x,y),3)
                    ,"lm"=paste0("y = ",sprintf(round(coef(lm(y~x)),2)[1],fmt="%1.2f")," + ",sprintf(round(coef(lm(y~x)),3)[2],fmt="%1.3f"),"x")
                    #,"lm_R2"=round(summary(lm(y~x))$r.squared,2)
                    )
              , by="dataset"]

knitr::kable(anscombe.summary)





```


## More illustrations on data patterns and correlation coefficients

- https://janhove.github.io/teaching/2016/11/21/what-correlations-look-like





# First plot

- Data: https://en.wikipedia.org/wiki/Iris_flower_data_set


```{r}
# load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load("data.table")
pacman::p_load("ggplot2")

# to data table [for data management](optional)
iris <- as.data.table(iris)

str(iris)

summary(iris)


# Scatterplot
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()

# show unique values of Species
unique(iris$Species)

# Show subset
iris[Species=="setosa"]

# Scatter plot Subset
ggplot(data=iris[Species=="setosa"], aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()

# Introduce color
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species))+
  geom_point()

# Boxplot
ggplot(data=iris, aes(x=Species, y=Sepal.Length))+
  geom_boxplot()  

# Reduce data set to first 5 Rows
iris[1:5,]

# Combine boxplot and 'scatterplot'
ggplot(data=iris[1:5,], aes(x=Species, y=Sepal.Length))+
  geom_boxplot()+
  geom_point()

# Boxplot for 5 Data points does not make much sense!

```



# Exercise 

## Part 1 - Scatter plot

- Open R/RStudio and load the packages "ggplot2" and "data.table"
- Create a plot comparing Petal.Width and Petal.Length using scatter plots for the entire iris data set 
- Create singe plots for each species
- Plot all data, use color to distinguish between species 
- Plot all data, use shapes to distinguish between species

## Part 2 - Box plot

- Compare all four numerical variables describing the different species using box plots.





# Additional resources

- https://en.wikipedia.org/wiki/Anscombe%27s_quartet
- https://style.tidyverse.org/ggplot2.html

#  Session Info

```{r}

sessionInfo()

```
