---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r substr((Sys.Date()),1,4)`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 002: First plot




# R and RStudio

## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

usePackage <- function(p) 	
{	
  if (!is.element(p, installed.packages()[,1]))	
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1])	
  require(p, character.only = TRUE)	
}	
	
usePackage("data.table")
usePackage("ggplot2")


```





# First plot

- Data: https://en.wikipedia.org/wiki/Iris_flower_data_set


```{r}
# load packages
library(data.table)
library(ggplot2)

# to data table [for data management](optional)
iris <- as.data.table(iris)

str(iris)

summary(iris)


# Scatterplot
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()

# show unique values of Species
unique(iris$Species)

# Show subset
iris[Species=="setosa"]

# Scatter plot Subset
ggplot(data=iris[Species=="setosa"], aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()

# Introduce color
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species))+
  geom_point()

# Boxplot
ggplot(data=iris, aes(x=Species, y=Sepal.Length))+
  geom_boxplot()  

# Reduce data set to first 5 Rows
iris[1:5,]

# Combine boxplot and 'scatterplot'
ggplot(data=iris[1:5,], aes(x=Species, y=Sepal.Length))+
  geom_boxplot()+
  geom_point()

# Boxplot for 5 Data points does not make much sense!

```



# Exercise 

## Part 1 - Scatter plot

- Open R/RStudio and load the packages "ggplot2" and "data.table"
- Create a plot comparing Petal.Width and Petal.Length using scatter plots for the entire iris data set 
- Create singe plots for each species
- Plot all data, use color to distinguish between species 
- Plot all data, use shapes to distinguish between species

## Part 2 - Box plot

- Compare all four numerical variables describing the different species using box plots.




#  Session Info

```{r}

sessionInfo()

```
