---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---
```{r echo=FALSE}

source("./src/custom.R")

```


# 002: A First - exploratory - plot


# 🎯 Learning objectives

::: {.callout-note icon=false}
## ggplot2 for plotting

**... understanding concepts:**

*  you know the **three essential components** for a `ggplot`    
*  you can follow the ggplot2 *Gramma of Graphics* to read plotting code sequentially  

**... tool usage:**

*   you can create a simple **exploratory** plot from a given data set  

:::



<br>

::: color-box
:::


# Why `ggplot2`?

- **reproducible results** (Script/Rmarkdown/Quarto) 
- when working in R it's the most comprehensive plotting environment 
- free and open source
- large community, well documented (problem solving) 
- very flexible *A Layered Grammar of Graphics*
  - syntax based on *Wilkinson, Leland. 2005. The Grammar of Graphics. 2nd ed. Statistics and Computing. Springer*. 
  - transferred to R mainly by Hadley Wickham (<http://hadley.nz/>, <https://ggplot2-book.org/>, <https://ggplot2.tidyverse.org/>)  
  - 'A Layered Grammar of Graphics. (Wickham, Hadley. 2009. 'A Layered Grammar of Graphics.' Journal of Computational and Graphical Statistics.) 
- `ggplot2` is also part of the `tidyverse` - alternative coding style with pipes `%>%` applicable






<br>

::: color-box
:::



# Prerequisities

- load `packages` and get a data  (`data.frame`, `tibble`)    
- data: [Iris flower data](https://en.wikipedia.org/wiki/Iris_flower_data_set)    


```{r, message=FALSE}
# load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse")

```



<br>

::: color-box
:::

# ggplot2 - Essential components

In the age of AI coding assistants, you will rarely need to write a ggplot2 script entirely from memory, but **you will have to act as an Editor**.

**However, to be a good editor, you must understand the language.** ggplot2 is built on the Layered Grammar of Graphics. Every plot is built using three essential components:

1.  **Data (`data`):** The R tabulated data (usually a `data.frame` or a `tibble`).    
2.  **Mapping (`mapping = aes()`):**  A so called *Aesthetic Mapping*, linking columns in your data to visual aesthetics (x, y, color, shape)...      
3.  **Geomery (`geom_...`):** The physical shapes drawn on the screen (points, lines, bars). (points, lines, boxplot, ...)?     



<br>

::: color-box
:::



## Know your data ... 

```{r}
# to tibble [for data management](optional)
iris <- as_tibble(iris)

iris
```


::: color-box 
::: 

<br>
<br>

## `ggplot(data=...)`   

```{r}
# 1. Only the data - empty plot
ggplot(data=iris)

```

## `mapping=aes(...)` (2)

```{r}
#2. data + mapping (aesthetics): grid without geometrical objects

ggplot(data = iris, 
       mapping = aes(x = Sepal.Length, y = Sepal.Width))

```

::: color-box 
::: 

<br>
<br>

## `+ geom_...` (3) = **First plot**  


```{r}
# 3. data + mapping + geometry (added geom layer with + sign)
# Frist scatterplot
ggplot(data = iris, 
       mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()
```

::: color-box 
::: 

<br>
<br>

## More global mapping

```{r}
ggplot(
  data = iris,
  aes(
    x = Sepal.Length,
    y = Sepal.Width,
    color = Species   # added color mapping
  )
) +
  geom_point()
```


::: color-box 
::: 

<br>
<br>

## Supply/modify data - tidyverse style `%>%`

```{r}
# data is *flowing* through the pipe into the plot
iris %>% 
  filter(Species == "setosa") %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) +
    geom_point()
```

::: color-box 
::: 

<br>
<br>

## Another geom_

```{r}
# Boxplot
ggplot(data=iris, aes(x=Species, y=Sepal.Length))+
  geom_boxplot()  
```


::: color-box 
::: 

<br>
<br>

### A gentle reminder on box plots


::: {style="font-size: 130%;"}
A boxplot for 5 (or even less) data points does not make much sense! 
:::

```{r}
# Reduce data set to first 5 Rows
iris[1:5,]

# Combine boxplot and 'scatterplot' (strip plot)
ggplot(data=iris[1:5,], aes(x=Species, y=Sepal.Length))+
  geom_boxplot()+
  geom_point(color="red")

```



::: color-box 
::: 

<br>
<br>

# Exercise 

::: {.callout-tip icon=false}
## 👉 Your first plots 

**Part 1 - Scatter plot**

- Goto your R script and load the `tidyverse` package and the `iris` data

```{r}
#| eval: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse")
iris <- as_tibble(iris)
```

- Create a plot comparing `Petal.Width` and `Petal.Length` using a scatter plot for the entire iris data set 
- Create separate plots for each species (use the *powerful* `copy&paste` workflow)
- Plot all data, use `color`s to distinguish between species 
- Plot all data, use `shape`s to distinguish between species

<br>

**Part 2 - Box plot**

- Compare all four numerical variables describing the different species using box plots.


:::

---  

::: {.callout-warning icon=false}
## Warning note on copy and past workflow  
 
> **Copy-and-paste is a powerful tool, but** you should avoid doing it more than twice. Repeating yourself in code is **dangerous** because it can easily lead to errors and inconsistencies.   
> 
>    
> Q: R for Data Science (2e) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund.   
> > <https://r4ds.hadley.nz/program.html>    
::: 
 


# Additional resources

- [ggplot tidyverse code style guide](https://style.tidyverse.org/ggplot2.html)   
- [R for Data Science (2e)](https://r4ds.hadley.nz/)   



#  Session Info

```{r}

sessionInfo()

```
