---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---
```{r echo=FALSE}

source("./src/custom.R")

```


# 003: Exploratory Data analysis with various plot types


# 🎯 Learning objectives

::: {.callout-note icon=false}
## Why and how - Data visualization

**... understanding concepts:**

* you understand why exploratory data analysis (EDA) is a fast, **iterative process**.
* you understand the different purposes of **basic plot types** (what to use when).  

**... tool usage:**

*   you can quickly produce standard plot types for different data combinations using `ggplot2` with minimal code.  

:::


::: color-box 
::: 



## Cheatsheet 

- [Data visualization with ggplot2 : : CHEATSHEET](https://rstudio.github.io/cheatsheets/html/data-visualization.html)

<br>
<br><br>

::: color-box 
::: 


<br><br>

# Prerequisites


```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse")
pacman::p_load("ggforce")   # geom_sina
```

::: color-box 
::: 

<br>
<br><br>

# Plot Types for EDA

## 1. Distributions (One Continuous Variable)

**Why:** To understand the shape, spread, and central tendency of your data, and to spot anomalies/outliers.  

**Where:** A first step when analyzing a numeric variable (e.g., age, size, temperature).

```{r}
# Histogram: Bins data into distinct groups to show frequencies
ggplot(data = iris, aes(x = Sepal.Length)) +
  geom_histogram(binwidth = 0.2
                , boundary=0       # or center = ...    | KNOW where your bins start!
                , color = "white"  # (optional) coloring the border with a fixed color
           )

# Density: A smoothed, continuous version of the histogram
ggplot(data = iris, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.5) # alpha adds transparency
```

::: color-box 
::: 

<br>
<br><br>

## 2. Comparisons (Categorical x Continuous)

**Why:** To compare the distribution of a numeric variable across different groups.  

**Where:** Comparing outcome measurements by treatment, year, or species.

```{r}
# Boxplot: Great for summarizing median, quartiles, and outliers quickly
ggplot(data = iris, aes(x = Species, y = Petal.Width)) +
  geom_boxplot()

# Jitter: Shows the actual raw data points to reveal sample sizes and clusters
ggplot(data = iris, aes(x = Species, y = Petal.Width)) +
  geom_jitter(width = 0.2, alpha = 0.6, height=0)   

# Violin: Shows the shape of the density distribution (combines density + boxplot ideas)
ggplot(data = iris, aes(x = Species, y = Petal.Width)) +
  geom_violin()

# Sina: *Jitter inside violin* Shows the actual raw data points AND the shape of the density distribution
ggplot(data = iris, aes(x = Species, y = Petal.Width)) +
  ggforce::geom_sina(jitter_y=FALSE)

```

<br><br>

---

❓ **Visual-Information**

What can you extract from the sina regarding the measurement resolution?

---

<br>
<br><br>

::: {.callout-warning icon=false}
## Warning note on geom_jitter + geom_sina

Make sure to **disable jittering** in the direction **that would change the measurement value** (usually *y*-direction). 

```{r}
geom_jitter(height=0)              # jitter height = 0
ggforce::geom_sina(jitter_y=FALSE) # disable jittering in y-direction
```



::: 
 

::: color-box 
::: 


::: {.callout-warning icon=false}
## Warning note on geom_violin + geom_density + geom_sina

Interpret with caution if there are fewer than approx. 50 observations in a group.

See **Exercise**

::: 


<br><br>

::: color-box 
::: 




## 3. Frequencies & Amounts (Categorical Variables)

**Why:** To see how many observations belong to a category, or to plot a pre-calculated summary statistic.  
**Where:** Survey responses, inventory/yield counts, ...

```{r}
# geom_bar implicitly COUNTS the number of rows per category
ggplot(data = iris, aes(x = Species)) +
  geom_bar()

# more intersting with differnt counts per species
set.seed(1)
iris %>% 
  slice_sample(n=75) %>%           #random sample 
  ggplot(aes(x = Species)) +
    geom_bar()


# geom_col plots exact VALUES provided in the dataset (e.g., summary statistics)
# Here, we calculate average highway mileage per car class
set.seed(1)
iris_summary <- iris %>%
  slice_sample(n = 75) %>%
  group_by(Species) %>%
  summarise(n_obs = n())

iris_summary

# reorder() is recommended to make bar plots easily readable
# IF no fixed order is required
ggplot(data = iris_summary, 
         aes(x = n_obs                        # easy rotation of the plot 
          , y = reorder(Species, n_obs))      # by changing x+y mapping
       ) +
  geom_col() 
```




<br><br>

::: color-box 
::: 



## 4. Relationships (Two Continuous Variables)

**Why:** To observe correlations, trends, or clusters between two numeric values.  

**Where:** Plotting height vs. weight, price vs. age, or temperature vs. yield, etc...  

```{r}
# Scatterplot
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point()

# Scatterplot separated by a third categorical variable
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()

```




## 5. Time-Series (Continuous Time x Continuous Value)

**Why:** To observe how a variable changes chronologically.  

**Where:** Weather data, stock prices, or sensor data (data logger) ...

```{r}
# built-in 'economics' dataset
# geom_line connects points sequentially by the x-axis
head(economics)

ggplot(data = economics, aes(x = date, y = unemploy)) +
  geom_line()
```


::: {.callout-warning icon=false}
## Missing Data in timeseries

```{r}
# MESS_DATUM  = Measurement Date
# TMK.Lufttemperatur = Daily Mean Air Temperature 2 m above ground 
# RSK.Niederschlagshoehe = Daily sum of precipitation (mm)

clim <- read_csv("clim_DWD_Geisenheim_current.csv",show_col_types = FALSE) %>%
            select(MESS_DATUM
                   ,TMK.Lufttemperatur
                   ,RSK.Niederschlagshoehe)

head(clim)

# Line plot
ggplot(data=clim, aes(x=MESS_DATUM, y=TMK.Lufttemperatur))+
  geom_line()

# 'introduce' missing data 
# Scatterplot - to show the GAP
ggplot(data=clim[-c(100:150),], aes(x=MESS_DATUM, y=TMK.Lufttemperatur))+
  geom_point()   

# Dot not plot timeseries data with 'removed' time stamps alone!
ggplot(data=clim[-c(100:150),], aes(x=MESS_DATUM, y=TMK.Lufttemperatur))+
  geom_line()

# Correclty hand missing data by setting/keeping values for missing times to NA
clim_NA <- clim 
clim_NA$TMK.Lufttemperatur[c(100:150)] <- NA
ggplot(data=clim_NA, aes(x=MESS_DATUM, y=TMK.Lufttemperatur))+
  geom_line()  

```

:::

<br>
<br><br>


::: color-box 
::: 


## 6. High-Density Data (Overplotting)
**Why:** When dealing with tens of thousands of rows, scatterplots become solid black blobs. You need to group points spatially.  
**Where:** Big data sets (e.g., flow cytometry, massive surveys, geographic tracking).

```{r}
# The 'diamonds' dataset has ~54,000 rows. A standard scatterplot is less informative here.
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05) # One solution: extreme transparency

# Alternative solution: 2D Binning (creates a heat map of counts)
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_bin2d(bins=30)

```


More alternatives as ggplot2 extensions ...

- e.g.  [ggpointdensity - geom_pointdensity(): A cross between a scatter plot and a 2D density plot.](https://github.com/LKremer/ggpointdensity)  


---



<br><br>


::: color-box 
::: 


# Exercise

::: {.callout-tip icon=false}
## 👉  Does sample size matter - Kernel density estimated *shapes*?

Density estimates (which create the shape of `geom_violin` or `geom_density`) require a decent amount of data to be accurate. If your sample size is too small, these plots can "hallucinate" shapes that aren't really there, or completely miss the true distribution. 

*(Concept inspired by the Mayo Clinic's [size_matters](https://rtools.mayo.edu/size_matters/) tool).*

**Your Task:**

*Variant A*

1. Go to <https://rtools.mayo.edu/size_matters/> and select *Simulator 1: The effect of sample size on normality tests and the accuracy of summary statistics*.
2. Change the values of 'n' in the boxes to view samples of different sizes.  
3. Repeat the same values via 'Draw a New Sample' to make sure your answers to the following/provided questions remain consistent ...

- *How many observations do you need to be confident that you can distinguish between the different data distributions?*   
-  *At what sample size does it become impossible for you to tell the difference between the data distributions?*    
- *What sample sizes do you usually work with? With samples of this size, can you tell the difference between the distributions?*    


---  
 
 
*Variant B*

1. Run the code below e.g. with the starting sample size of `n <- 3`.    
2. Look at the plot: Can you clearly see which distribution is Normal, Bimodal, or Skewed based on the violin shape alone?    
3. Change `n` to **10**, **20**, **50**, **100**, and **1000** and re-run the code. Watch how the violin shapes stabilize.     
4. Answer the questions above ... 


### Simulation Code

```{r}
#| fig-width: 7
#| fig-height: 5


# 1. CHANGE THIS NUMBER: Try 10, 20, 50, 100, 500, 1000 ... 
n <- 3 


# 2. Creat ground truth data from three known underlying distributions
n_truth = 1e4
ground_truth <- tibble(
  Normal  = rnorm(n_truth, mean = 0, sd = 1),
  Bimodal = ifelse(runif(n_truth) > 0.5, rnorm(n_truth, -3, 1), rnorm(n_truth, 3, 1)),
  Skewed  = rlnorm(n_truth, meanlog = 0, sdlog = 0.8)
) |> 
  # Reshape data for ggplot2
  pivot_longer(cols = everything(), names_to = "Distribution", values_to = "Value")


# 3. Simulate data with lower 'n'
sim_data <- tibble(
  Normal  = rnorm(n, mean = 0, sd = 1),
  Bimodal = ifelse(runif(n) > 0.5, rnorm(n, -3, 1), rnorm(n, 3, 1)),
  Skewed  = rlnorm(n, meanlog = 0, sdlog = 0.8)
) |> 
  # Reshape data for ggplot2
  pivot_longer(cols = everything(), names_to = "Distribution", values_to = "Value")


# 3. Plot the estimated shapes (violin) vs. actual data (points)
ggplot(sim_data, aes(x = Distribution, y = Value, fill = Distribution)) +
  geom_violin(data=ground_truth, fill=NA, color = "black", trim = FALSE) +    # The ground truth shape
  geom_violin(alpha = 0.4, color = NA, trim = FALSE) +               # The estimated shape
  ggforce::geom_sina(alpha = 0.7) +         # The raw data points
  theme_minimal() +
  labs(
    title = paste("Sample Size (n) =", n, "per group"),
    subtitle = "Do the violins accurately show normal, bimodal, and skewed shapes?",
    x = NULL, 
    y = "Value"
  ) +
  theme(legend.position = "none")


```

:::


<br><br>


::: color-box 
::: 


## Bonus: Dot and Lollipop instead of bar plot

- Dataset: USArrests - arrests per 100.000 residents per State and type of crime

```{r}
data(USArrests)
USArrests <- as_tibble(USArrests,rownames = "State")
USArrests

# bar plot
ggplot(USArrests,aes(x = State, y = Assault)) +
    geom_col() 

# rotate bar
ggplot(USArrests,aes(y = State, x = Assault)) +
    geom_col()
 
# Reorder States 
# bar    
ggplot(USArrests,aes(y = reorder(x=State,X=Assault,FUN=max), x = Assault)) +
    geom_col()+
    scale_x_continuous(limits=c(0,NA)                       # include zero
                      ,expand = expansion(mult = c(0, .1))) # expand right 


# dot plot
ggplot(USArrests,aes(y = reorder(x=State,X=Assault,FUN=max), x = Assault)) +
    geom_point()+
    scale_x_continuous(limits=c(0,NA)                       # include zero
                      ,expand = expansion(mult = c(0, .1))) # expand right  

# lollipop
ggplot(USArrests,aes(y = reorder(x=State,X=Assault,FUN=max), x = Assault)) +
    geom_pointrange(aes(xmin=0,xmax=Assault))+
    scale_x_continuous(limits=c(0,NA)                       # include zero
                      ,expand = expansion(mult = c(0, .1))) # expand right  


```



<br><br>


::: color-box 
::: 


# Bonus: Download `rdwd` weather data

```{r, eval=FALSE}
# load packages
pacman::p_load("rdwd")
#to update to latest version
updateRdwd()

##########################################################################################
# DO NOT EXECUTE CODE IN BINDER !
##########################################################################################
# #Select data from the DWD CDC FTP Server
# #set current=TRUE - to check current FTP Server Index to gather up-to-date file information
link <- selectDWD("Geisenheim", res="daily", var="kl", per="recent", current=FALSE)
#Download data from the DWD CDC FTP Server
file <- dataDWD(link, read=FALSE, dir="DWDdata", quiet=TRUE, force=NA)
#Process data from the DWD CDC FTP Server to a data.frame
clim <- readDWD(file, varnames=TRUE, fread=TRUE)

# # Write out for reuse
write_csv(clim[1:365,],"clim_DWD_Geisenheim_current.csv")
```


<br><br>


::: color-box 
::: 


# Additional resources


- [ggpointdensity](https://github.com/LKremer/ggpointdensity)
- [rdwd](https://cran.r-project.org/web/packages/rdwd/vignettes/rdwd.html)



#  Session Info

```{r}

sessionInfo()

```
