---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "2020"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 004: ggplot2 layers

## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

usePackage <- function(p)   
{ 
  if (!is.element(p, installed.packages()[,1])) 
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1]) 
  require(p, character.only = TRUE) 
} 
  
usePackage("data.table")
usePackage("ggplot2")

```


# Why ggplot2?

- when working in R it's the most comprehensive plotting environment 
- free and open source
- large community (problem solving) 
- reproducible results (R-script/project) 
- very flexible ***A Layered Grammar of Graphics*** 
  - syntax based on ***Wilkinson, Leland. 2005. The Grammar of Graphics. 2nd ed. Statistics and Computing. Springer***. 
  - transferred to R mainly by Hadley Wickham (http://hadley.nz/, https://ggplot2-book.org/, https://ggplot2.tidyverse.org/)  
  - 'A Layered Grammar of Graphics. (Wickham, Hadley. 2009. 'A Layered Grammar of Graphics.' Journal of Computational and Graphical Statistics.) 
- `ggplot2` is also part of the `tidyverse` - alternative coding style with pipes `%>%` 



## ggplot2 syntax / Layered Grammar of Graphics

A basic overview:

- build plot with different layers
- first layer usually contains the data
- every layer has 'aesthetic mappings' to assign variables from the data to plotting objects (x, y, color, shape, ...)
- new layers are 'added' with a `+`
- basic graphical layers are called (`geom_...`) 
- other layers (`stat_...`) can be used for more specific statistical transformations 
- and there are layers for modifying the appearance (coordinate system, axis, fonts, legends, etc.) 



### graphical Layers -- `geom`s

Examples:

- `geom_point` - creates points
- `geom_line` - ... lines
- `geom_histogram` - ... a histogram
- `geom_boxplot` - ... boxplot 
- ...

- see cheat sheet for more options https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf


### 'Aesthetic Mappings' `aes`

- assign 'what is what' with ('mappings')
- besides basic mappings, e.g. which variable should be plotted on x and y axis (`x=...`, `y=...`), you can also 'map'
- colors (`color`,`fill`)
- shapes (`shape`,`linetype`)
- sizes (`size`)

Not all `geom`s are compatible with all the 'aesthetic mappings'.
You cannot assign a linetype to a point representation.



# Combine geoms

- add new layers with a `+`

```{r}
library("ggplot2")
library("data.table")

##########################################################################################
# Read file
##########################################################################################
clim <- fread("clim_DWD_Geisenheim_current.csv")
clim[,"MESS_DATUM":=as.POSIXct(MESS_DATUM)] 

##########################################################################################

# combine plots
ggplot(data=clim[-c(100:150),], aes(x=MESS_DATUM, y=TMK.Lufttemperatur))+ # Init Data
  geom_point()+ # Layer 1
  geom_line()   # Layer 2 [above Layer 1]


```


# Regression line

- "Trendlinie" 

```{r}

# regression line across all data
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()+
  geom_smooth(method="lm")   #linear model

# Without specification of linear model - automatic model selection
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()+
  geom_smooth()

# LM: Linear regression model  
# LOESS (locally estimated scatterplot smoothing) 
# Alternative: generalized additive model (GAM) 

# Regression line for subsets [grouped by color = species] 
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species))+
  geom_point()+
  geom_smooth(method="lm")  #linear model


```


# Boxplot + jitter

- combine plots + options
- options: `?geom_jitter`; `?geom_dotplot` ... etc.

```{r}

# jitter is more appropriate for more and real continuous data [here: 0.1 cm 'categories']
ggplot(data=iris, aes(x=Species,y=Petal.Width))+
  geom_boxplot()+
  geom_jitter( height = 0)   #do not allow random positions in y-direction

# better alternative dotplot
ggplot(data=iris, aes(x=Species,y=Petal.Width))+
  geom_boxplot()+
  geom_dotplot(binaxis='y', stackdir='center', binwidth = 0.1,dotsize=0.25)


```




# Plot objects

- save plots to variables
- e.g. to add layers step by step
- modify appearance afterwards


```{r}
# Save a plot object into a variable - without displaying the plot
p0 <- ggplot(iris,aes(x=Petal.Length, y= Petal.Width))+
        geom_point()

# display plot by calling the variable
p0

# add layer to p0 and display plot
p0 +
  geom_smooth(method="lm")


# save new plot to another variable
p1 <- p0 +
          geom_smooth(method="lm")

# output p1 
p1 

# additional layers for appearance modification, e.g. 'themes' [details follow soon]
p0 +
  theme_bw()


# Store only the data initialization layer to generate different types of plots 
p00 <- ggplot(iris,aes(x=Species, y=Petal.Width))

p00 + geom_boxplot()

p00 + geom_violin()

```



# Additional resources


- https://en.wikipedia.org/wiki/Local_regression
- https://en.wikipedia.org/wiki/Generalized_additive_model
- http://www.sthda.com/english/wiki/ggplot2-dot-plot-quick-start-guide-r-software-and-data-visualization

#  Session Info

```{r}

sessionInfo()

```
