---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---
```{r echo=FALSE}

source("./src/custom.R")

```


# 004: ggplot2 layers


# 🎯 Learning objectives

::: {.callout-note icon=false}
## Layered graphics

**... understanding concepts:**


* you understand why and where **exploratory graphs benefit from layering**, i.e. combining multiple *geoms* in a `ggplot`   
* you understand the logic in a the `ggplot` code with multiple geometric layers  
* you know of the possibility to store a `ggplot` in a variable (for modification, export, etc.)

**... tool usage:**

*   you can add multiple geoms to a single plot in a requested order      

:::


<br><br>

::: color-box 
::: 





# Prerequisites


```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse")
pacman::p_load("ggforce")   # geom_sina
```

<br><br>

::: color-box 
::: 


# Combine compatible geoms  

- add new layers with a `+`


<br><br>

::: color-box 
::: 


## Boxplot + Violing + Raw data 


::: {style="font-size: 120%;"}

- *the egg-laying, wool-milk-sow* ('die Eierlegende Wollmilchsau') *of univariate data distribution exploration*
:::

```{r}

ggplot(iris,aes(x=Species,y=Petal.Length))+
    geom_violin()+                         # (optional) supports sina-shape
    ggforce::geom_sina(jitter_y=FALSE)+    # deactivate value-jitter
    geom_boxplot(width=0.1          # adjust to place inside the violin
                ,alpha=0.5          # adjust to see the raw data 
                ,outliers = FALSE   # no 'duplicate' data points
                )

```

<br><br>

::: color-box 
::: 

# Regression line

- The famous *Excel*-***Trendlinie***

```{r}

# regression line across all data
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()+
  geom_smooth(method="lm")   #linear model

# Without specification of linear model - automatic model selection
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()+
  geom_smooth()

# Methods
# "lm":   Linear regression model [stats::lm]
# "glm":  Generalized linear model [MASS::rlm]
# "loess" (Local Polynomial Regression Fitting) [stats::loess]
# "gam":  generalized additive model (GAM) [mgcv::gam]

# Regression line for subsets [grouped by color = Species] 
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species))+
  geom_point()+
  geom_smooth(method="lm")  #linear model


```


<br><br>

::: color-box 
::: 


# Lineplot + data points

## Missing data spotting

```{r, echo=FALSE}

clim <- read_csv("clim_DWD_Geisenheim_current.csv",show_col_types = FALSE) %>%
            select(MESS_DATUM
                   ,TMK.Lufttemperatur
                   ,RSK.Niederschlagshoehe)

```


```{r}
# Scatterplot + (bad) Line
ggplot(data=clim[-c(100:150),], aes(x=MESS_DATUM, y=TMK.Lufttemperatur))+
  geom_point()   +
  geom_line()
```

<br><br>

::: color-box 
::: 

## Irregular time step indication

- Growth curves

```{r}
# The Orange data frame has 35 rows and 
# 3 columns of records of the 
# growth of orange trees.
orange <- as_tibble(datasets::Orange)    #growth over time


ggplot(orange, aes(x=age,y=circumference,color=Tree))+
  geom_line(alpha=0.5)+
  geom_point(size=2)


```

<br><br>
<br><br>

::: {.callout-warning icon=false}
## Why these strange ordering of the legend?

```{r}
str(orange$Tree)

```

(orderd) factors / factor levels are respected in `ggplot`
if no levels are given (simple character data) alphabetical/chronological ordering is applied.
:::

<br><br>

::: color-box 
::: 






# Plot objects

- save plots to variables
- e.g. to add layers step by step
- modify appearance afterwards


```{r}
# Save a plot object into a variable - without displaying the plot
p0 <- ggplot(iris,aes(x=Petal.Length, y= Petal.Width))+
        geom_point()

# display plot by calling the variable
p0

# add layer to p0 and display plot
p0 +
  geom_smooth(method="lm")


# save new plot to another variable
p1 <- p0 +
          geom_smooth(method="lm")

# output p1 
p1 

# additional layers for appearance modification, e.g. 'themes' [details follow soon]
p0 +
  theme_bw()


# Store only the data initialization layer to generate different types of plots 
p00 <- ggplot(iris,aes(x=Species, y=Petal.Width))

p00 + geom_boxplot()

p00 + geom_violin()

```


<br><br>

::: color-box 
::: 

# Exercise

::: {.callout-tip icon=false}
## 👉  *Never trust summary statistics alone*?

<br><br>

- What summary statistics is misleading here when comparing the left (A) and right (B) plot variants?


```{r}
#| echo: false
#| message: false
pacman::p_load("patchwork")

# regression line across all data
p1 <- ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()+
  geom_smooth(method="lm")+   #linear model
  coord_fixed()

# Regression line for subsets [grouped by color = Species] 
p2 <- ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species))+
  geom_point()+
  geom_smooth(method="lm")+  #linear model
  coord_fixed()

(p1 | p2) + plot_annotation(tag_levels = "A")

```


:::

<br>
*Bonus*: Code to reproduce these plots


```{r}
#| echo: true
#| eval: false
pacman::p_load("ggplot2")
pacman::p_load("patchwork")

# regression line across all data
p1 <- ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width))+
  geom_point()+
  geom_smooth(method="lm")+  
  coord_fixed()

# Regression line for subsets [grouped by color = Species] 
p2 <- ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species))+
  geom_point()+
  geom_smooth(method="lm")+ 
  coord_fixed()

(p1 | p2) + plot_annotation(tag_levels = "A")


```


<br><br>

::: color-box 
::: 


# Additional resources

- <https://ggplot2.tidyverse.org/reference/geom_smooth.html>



#  Session Info

```{r}

sessionInfo()

```