---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r substr((Sys.Date()),1,4)`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 005: Colors, shapes and linetypes

- `aes()` - aesthetics not only assign the data (variables) to the axis 
- can also be used to associate colors, shapes etc. with variables




## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

usePackage <- function(p)   
{ 
  if (!is.element(p, installed.packages()[,1])) 
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1]) 
  require(p, character.only = TRUE) 
} 
  
usePackage("data.table")
usePackage("ggplot2")

```


# Colors for points and lines - `color` 

## `color` grouping for discrete variables

- `color` can act as a grouping variable

```{r}
library("data.table")
library("ggplot2")

iris <- as.data.table(iris)


# global color association
ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width,color=Species))+
  geom_point()+
  geom_smooth(method="lm")

# local color association
ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_smooth(method="lm")


```


## `color` and numerical variables

- color bar

```{r}

ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Petal.Length))


```


# `shape`

## `shape` grouping for discrete variables


- combine plots + options
- options: `?geom_jitter`; `?geom_dotplot` ... etc.

```{r}

ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width, shape=Species))+
  geom_point()


# local
ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(shape=Species))+
  geom_smooth(method="lm")

# global association
ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width,shape=Species))+
  geom_point()+
  geom_smooth(method="lm")

```


## Grouping: `color` + `shape`

```{r}
ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width,shape=Species,color=Species))+
  geom_point()+
  geom_smooth(method="lm")

```


## *List of available shapes* 

- https://ggplot2.tidyverse.org/reference/scale_shape.html 
- filled red shapes (21-25) also allow a `fill` color besides the `color` assignment [=border color] 

```{r}

df_shapes <- data.table(shape = 0:25)
ggplot(df_shapes, aes(0, 0, shape = shape)) +
  geom_point(aes(shape = shape), size = 5, fill = 'red') +
  scale_shape_identity() +
  facet_wrap(~shape) +
  theme_void()

```

# `linetype`


```{r}
# re-assign to data.table 
economics_long <- as.data.table(economics_long)  

str(economics_long)

# data in 'long format'
economics_long

# grouping by linetype
ggplot(economics_long, aes(x=date, y=value01,linetype=variable)) + 
  geom_line()

# grouping by color
ggplot(economics_long, aes(x=date, y=value01,color=variable)) + 
  geom_line()

# both
ggplot(economics_long, aes(x=date, y=value01,color=variable, linetype=variable)) + 
  geom_line()


```

- ***choose wisely*** - `linetype` harder to distinguish than `color` -- when colors are limited to high contrasts.


# fill-colors: `fill`

- for areas: box plot, histogram, etc.

```{r}

# fill
ggplot(data=iris, aes(x=Species,y=Sepal.Length))+
  geom_boxplot(aes(fill=Species))

# color
ggplot(data=iris, aes(x=Species,y=Sepal.Length))+
  geom_boxplot(aes(color=Species))

```


- for points 
- special shapes needed (21-25)

```{r}
# color and default shapes
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))

# not working 
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(fill=Species))

# assign fixed shapes that allow filling
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(fill=Species), shape=21)             # single shape as an "option"

# not working out of the box- shape by species
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(fill=Species, shape=Species))

# need to assign shapes separately
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(fill=Species, shape=Species))+
  scale_shape_manual(values=21:23)              # manual shape assignment layer

```


-  `scale_...`-layers are for modulating all kind of 'scales', e.g. including axis, colors, shapes, linetypes ...
- standard border color = 'black' 





# Best Practices

## Find the red dot

- based on https://socviz.co/lookatdata.html (Figure 1.18)

```{r echo=FALSE, message=FALSE, fig.cap="*Searching for the red dot becomes progressively harder.*"}

set.seed(1)
d20 <- data.table(x=sample(x = seq(0,30,by=0.01), size=20),
                  y=sample(x = seq(0,100,by=0.01), size=20),
                  color=c(rep("B",times=19),"A"),
                  shape=c("X",rep("Y",times=19))
                  )

d100 <- data.table(x=sample(x = seq(0,30,by=0.01), size=100),
                  y=sample(x = seq(0,100,by=0.01), size=100),
                  color1=c(rep("B",times=99),"A"),
                  shape1=c(rep("Y",times=99),"X"),
                  color2=c(rep("A",times=51),rep("B",times=49)),
                  shape2=c(rep("Y",times=50),rep("X",times=50))
                  )


p1 <- ggplot(d20,aes(x=x,y=y,color=color))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe, N=20")+
  xlim(0,30)+ylim(0,100)

p2 <- ggplot(d100,aes(x=x,y=y,color=color1))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe, N=100")+
  xlim(0,30)+ylim(0,100)

p3 <- ggplot(d20,aes(x=x,y=y,shape=shape))+
  geom_point(color="#F8766D")+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Form, N=20")+
  xlim(0,30)+ylim(0,100)

p4 <- ggplot(d100,aes(x=x,y=y,shape=shape1))+
  geom_point(color="#F8766D")+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Form, N=100")+
  xlim(0,30)+ylim(0,100)


p5 <- ggplot(d100,aes(x=x,y=y,color=color2,shape=shape2))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe+Form, N=100")+
  xlim(0,30)+ylim(0,100)


library(patchwork)

p1|p2|p3|p4|p5

```


## Find the red dot 2


```{r echo=FALSE, message=FALSE}

p5 <- ggplot(d100,aes(x=x,y=y,color=color2,shape=shape2))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe+Form, N=100")+
  xlim(0,30)+ylim(0,100)

p6 <- ggplot(d100,aes(x=x,y=y,color=paste(color2,shape2)))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("3 Farben, N=100")+
  xlim(0,30)+ylim(0,100)

p5|p6
```



## `alpha` - for overlaps and appearance 

- parameter `alpha` is limited between 0 - 1 
- `alpha=0` is *invisible*/*transparent* object  
- `alpha=1` standard = not-transparent 



```{r}

library(ggplot2)
library(data.table)
#generate overlapping scatterplot data by sampling from a normal distribution for x+y
set.seed(1) #fix random sampling with 'seed'
dRandom <- data.table( 
                        x=rnorm(n=1000,mean=0,sd=0.5),
                        y=rnorm(n=1000,mean=0,sd=0.5)
                      )

str(dRandom)


#default
ggplot(dRandom, aes(x=x, y=y))+
  geom_point()


#alpha=0.5 fix
ggplot(dRandom, aes(x=x, y=y))+
  geom_point(alpha=0.5)


#alpha depending on x-values'
ggplot(dRandom, aes(x=x, y=y))+
  geom_point(aes(alpha=x))

```



# Color palettes


## Manual color selection

* `scale_color_manual`
* `scale_fill_manual`

- color names: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

```{r}
# With color names


#internal definition
ggplot(data=iris, aes(x=Sepal.Length,y=Sepal.Width))+
  geom_point(aes(color=Species))+
  scale_color_manual(values=c("red","orange","blue"))

#external (global) variable
cols <- c("red","orange","blue")

ggplot(data=iris, aes(x=Sepal.Length,y=Sepal.Width))+
  geom_point(aes(color=Species))+
  scale_color_manual(values=cols)

```



### RGB or HEX

- RGB https://en.wikipedia.org/wiki/RGB_color_model
- HEX https://www.color-hex.com


```{r}

#create 3 colors by RGB definition
hgu_petrol      <- rgb(0, 108, 125, maxColorValue = 255)
hgu_lightgray   <- rgb(197, 198, 198, maxColorValue = 255)
hgu_lightgreen  <- rgb(174, 195, 11, maxColorValue = 255)


# Combine colors in a vector for plotting
hgu_3cols <- c(hgu_petrol,hgu_lightgray,hgu_lightgreen)


# Use colors in ggplot
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_color_manual(values=hgu_3cols)


# improved visibility of light colors with fill and shape 
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
      geom_point(aes(fill=Species), shape=21)+
      scale_fill_manual(values=hgu_3cols)+
      theme_bw()


```

```{r}

# 3 HEX-Code Colors
cols <- c("#ff0000","#ffa500","#0000ff")

ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_color_manual(values=cols)

```


- `scale_color_identity` could be used when there is a color-code column in the dataset 
- https://ggplot2.tidyverse.org/reference/scale_identity.html




## Predefined color palettes

- https://ggplot2.tidyverse.org/reference/scale_grey.html


### ggplot2-Optionen

A small selection: 
> `scale_colour_grey`, `scale_colour_brewer`, `scale_colour_hue`, `scale_colour_viridis_d`
... etc.

- all palettes exist for `_color_` and `_fill_`


```{r}
# grayscale
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_colour_grey()

#default color scale
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_colour_hue()

# modification of h range -- check ?scale_colour_hue
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_colour_hue(h=c(0,50))
```

```{r}

ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_colour_brewer()

```


- `scale_colour_brewer()` palettes

```{r  echo=FALSE}
par(cex=0.5)
RColorBrewer::display.brewer.all()+theme_void(base_size = 8)

```

... und reduziert auf möglichst *colorblindFriendly* Paletten:

```{r  echo=FALSE}

par(cex=0.5)
RColorBrewer::display.brewer.all(colorblindFriendly = TRUE)

```


```{r}

ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_color_brewer(palette="Set2")

```


- https://ggplot2.tidyverse.org/reference/scale_viridis.html

```{r}

ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_colour_viridis_d()


# improve visibility with fill+shape for light colors
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
     geom_point(aes(fill=Species), shape=21)+
     scale_fill_viridis_d(option="C")

```


### more packages with color palettes


* `ggsci` -- https://nanx.me/ggsci/
* `wesanderson` -- https://github.com/karthik/wesanderson
* `unikn` -- https://cran.r-project.org/web/packages/unikn/vignettes/colors.html
* ...

Longer list: https://github.com/EmilHvitfeldt/r-color-palettes



# Manual `shape`-selection

```{r}

ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(shape=Species))+
  scale_shape_manual(values=c(2,3,4))

```

```{r echo=FALSE}

df_shapes <- data.table(shape = 0:25)
ggplot(df_shapes, aes(0, 0, shape = shape)) +
  geom_point(aes(shape = shape), size = 5, fill = 'red') +
  scale_shape_identity() +
  facet_wrap(~shape) +
  theme_void()


```


# Manual `linetype`-selection


```{r}

economics_long <- as.data.table(economics_long)
economics_long_red <- economics_long[variable%in%c("pop","unemploy")]

ggplot(economics_long_red, aes(x=date, y=value01)) + 
  geom_line(aes(linetype=variable))+
  scale_linetype_manual(values=c(2,3))

ggplot(economics_long_red, aes(x=date, y=value01)) + 
  geom_line(aes(linetype=variable))+
  scale_linetype_manual(values=c("dotted","dashed"))


```

```{r echo=FALSE}
#from ggpubr::show_line_types()

lt <- c("blank", "solid", "dashed", "dotted", "dotdash",  "longdash", "twodash")

d <- data.frame(lt = factor(lt, levels = lt))

ggplot() + scale_x_continuous(name = "", limits = c(0, 1), 
    breaks = NULL) + scale_linetype_identity() + geom_segment(data = d, 
    mapping = aes(x = 0, xend = 1, y = lt, yend = lt, linetype = lt)) + 
    labs(title = "Line types available in R", y = "") + theme(axis.text.y = element_text(face = "bold", 
    color = "black"))+
    theme_minimal()+
    scale_y_discrete(labels=paste(0:6,lt,sep=' = '))

```




# Exercise

- Play around with some options from the latest input 
- e.g. colors, bar plots, etc...
- ...



# Additional resources


- https://socviz.co/lookatdata.html
- https://en.wikipedia.org/wiki/Alpha_compositing
- https://en.wikipedia.org/wiki/Web_colors#Hex_triplet
- https://ggplot2.tidyverse.org/articles/ggplot2-specs.html
- https://ggplot2.tidyverse.org/reference/aes_linetype_size_shape.html
- https://www.rdocumentation.org/packages/ggthemes/versions/3.5.0/topics/scale_colour_few
- https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
- https://nanx.me/ggsci/
- https://lisacharlotterost.de/2016/04/22/Colors-for-DataVis/
- https://socviz.co/workgeoms.html

#  Session Info

```{r}

sessionInfo()

```
