---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---
```{r echo=FALSE}

source("./src/custom.R")

```


# 005: Colors, shapes and linetypes



# 🎯 Learning objectives

::: {.callout-note icon=false}
## Layered graphics

**... understanding concepts:**


* you know you can use `aes()` not only to assign the data (variables) to the axis, but also to **associate colors, shapes etc. with variables** from your  
* you understand the logic behind the **automatic legend creation** for such mappings    
* you know *best practices* in the color, shape and linetype selection for **clear and accessible** figure creation (beyond exploratory figures)   

**... tool usage:**

* you can **add and adjust** the color, shape and linetype of your geometric objects in a `ggplot`
*  you can **choose perceptually appropriate** color palettes for continuous, diverging, and categorical data.
*  you can **customize scales** (e.g., `scale_color_manual()`, `scale_shape_manual()`) to match publication standards.


:::


<br><br>

::: color-box 
::: 





# Prerequisites


```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse")
pacman::p_load("scico")   # scientific color palettes
```


<br><br>

::: color-box 
::: 


# Aesthetic Mappings in `aes()` 

The core idea in ggplot2 is the **grammar of graphics**: every visual property of a geometric object can be either *fixed* (set outside `aes()`) or *mapped* to a variable (set inside `aes()`).

| Property   | Fixed (outside `aes()`)       | Mapped (inside `aes()`)                    |
|------------|-------------------------------|--------------------------------------------|
| Color      | `color = "steelblue"`         | `aes(color = Species)`                     |
| Fill       | `fill = "tomato"`             | `aes(fill = Species)`                      |
| Shape      | `shape = 17`                  | `aes(shape = Species)`                     |
| Linetype   | `linetype = "dashed"`         | `aes(linetype = drv)`                      |
| Size       | `size = 3`                    | `aes(size = body_wt)`                      |
| Alpha      | `alpha = 0.4`                 | `aes(alpha = carat)`                       |

: Summary of aesthetic properties 


<br><br><br>

## Fixed vs. Mapped Aesthetics

```{r fixed-vs-mapped}
# Fixed
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(color = "steelblue")
             
# Mapped
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() 

```


::: {.callout-important}
## Key Rule
When an aesthetic is inside `aes()`, **ggplot2 automatically creates a legend** and maps the levels of the variable to distinct visual values (colors, shapes, etc.). When it is outside `aes()`, the value is applied uniformly to *all* observations.

**Thus, We usually want to use the mapping abilities**

:::

<br><br>

::: color-box 
::: 




# Global versus local 'mapping' 

- color (or shape) also act as grouping variables


```{r}

# global color association
ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width,color=Species))+
  geom_point()+
  geom_smooth(method="lm")

# local color assignment
ggplot(data=iris, aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_smooth(method="lm")


```



<br><br>

::: color-box 
::: 




#  `color`- Colors for points and lines

## Categorical Data: Qualitative Palettes 

Qualitative palettes assign **maximally distinct** hues to unordered categories. 
The default ggplot2 palette is evenly spaced on the color wheel - fine for exploration, but can be improved.

```{r color-categorical}
pacman::p_load("patchwork") # side by side plot

p_base <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  labs(title = "ggplot2 standard")

p_base

p1 <- p_base + 
  scale_color_brewer(palette = "Set2") + 
  labs(title = "ColorBrewer Set2")

p1

p2 <- p_base + 
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73")) + 
  labs(title = "Wong (colorblind-safe)")

p2

p3 <- p_base + 
  scale_color_viridis_d(option = "plasma") + 
  labs(title = "Viridis plasma (discrete)")

p3

p4 <- p_base + 
  scale_color_grey()

p4


```


<br><br>

- [Wong, B. Points of view: Color blindness. Nat Methods 8, 441 (2011). https://doi.org/10.1038/nmeth.1618](https://www.nature.com/articles/nmeth.1618)




<br><br>

::: color-box 
::: 




## Continuous Data: Sequential & Diverging Palettes

For **ordered** or **continuous** variables, use palettes that encode magnitude through lightness or saturation.

```{r color-continuous}
p_base <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Petal.Width)) +
            geom_point()
p_base

p_base +
  scale_color_viridis_c(option = "viridis") 

p_base +
  scico::scale_color_scico(palette="batlow",end=0.9) 



```




<br><br>

::: color-box 
::: 




# `shape`

## The Shape Scale

ggplot2 maps categorical variables to **25 built-in shapes** (0–25). Shapes 0–14 are outlines (controlled by `color`); shapes 15–20 are filled (controlled by `color`); shapes 21–25 are filled with a separate border (`fill` + `color`).

```{r shape-reference}
#| fig-cap: "All 25 built-in ggplot2 point shapes."
#| fig-height: 3
#| echo: false

df_shapes <- tibble(shape = 0:25)
ggplot(df_shapes, aes(0, 0, shape = shape)) +
  geom_point(aes(shape = shape), size = 5, fill = 'red') +
  scale_shape_identity() +
  facet_wrap(~shape) +
  theme_void()
```


- <https://ggplot2.tidyverse.org/reference/scale_shape.html>
- filled red shapes (21-25) also allow a `fill` color besides the `color` assignment [=border color] 



<br><br>

::: color-box 
::: 




## `shape` mapping

```{r}
p_base <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length
                           , shape = Species
                           , color=Species)) +
  geom_point() 

# Unified Legend of shape and color !
p_base

p1 <- p_base + 
  scale_shape_manual(values=c(15,16,18))

p1

p2 <- p_base + 
   scale_shape_manual(values=c(21,22,23))

p2

# fill
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, shape = Species,fill=Species)) +
  geom_point(color="black")+               # fixed mapping = all borders are black
  scale_shape_manual(values=c(21,22,15))   # 15=soild shape = not 'fillable'

```




<br><br>

::: color-box 
::: 






# fill-colors: `fill`

- for areas: box plot, histogram, etc.

```{r}

# fill
ggplot(data=iris, aes(x=Species,y=Sepal.Length))+
  geom_boxplot(aes(fill=Species))

# color
ggplot(data=iris, aes(x=Species,y=Sepal.Length))+
  geom_boxplot(aes(color=Species))

```


- for points 
- special shapes needed (21-25)

```{r}
# color and default shapes
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))

# not working 
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(fill=Species))

# assign fixed shapes that allow filling
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(fill=Species), shape=21)             # single shape as an "option"

# need to assign shapes separately
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(fill=Species, shape=Species))+
  scale_shape_manual(values=21:23)              # manual shape assignment layer

```


-  `scale_...`-layers are for modulating all kind of 'scales', e.g. including axis, colors, shapes, linetypes ...
- standard border color = 'black' 




<br><br>

::: color-box 
::: 





# `linetype`

## Available Linetypes

ggplot2 supports six named linetypes: `"solid"`, `"dashed"`, `"dotted"`, `"dotdash"`, `"longdash"`, `"twodash"`.

```{r linetype-reference}
#| fig-cap: "Six named linetypes available in ggplot2."
#| fig-height: 2.8

lt_df <- data.frame(
  y = 6:1,
  label = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash"),
  lt    = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash")
)

ggplot(lt_df) +
  geom_segment(aes(x = 0, xend = 1, y = y, yend = y, linetype = lt), linewidth = 1.2) +
  geom_text(aes(x = 1.03, y = y, label = label), hjust = 0, size = 3.5) +
  scale_linetype_identity() +
  xlim(0, 1.7) +
  theme_void() +
  labs(title = "Named linetypes")
```




<br><br>

::: color-box 
::: 




## Assign linetypes


```{r}
# re-assign to tibble 
economics_long <- as_tibble(economics_long)  

str(economics_long)

# data in 'long format'
economics_long


# no grouping 
ggplot(economics_long, aes(x=date, y=value01)) + 
  geom_line()

# grouping by 'group'
ggplot(economics_long, aes(x=date, y=value01,group=variable)) + 
  geom_line()

# grouping by linetype
ggplot(economics_long, aes(x=date, y=value01,linetype=variable)) + 
  geom_line()

# grouping by color
ggplot(economics_long, aes(x=date, y=value01,color=variable)) + 
  geom_line()

# both
ggplot(economics_long, aes(x=date, y=value01,color=variable, linetype=variable)) + 
  geom_line()


```




- ***NOTE*** - `linetype` is harder to distinguish than `color, when colors are limited to high contrast colors.



<br><br>

::: color-box 
::: 


## Manual `linetype`-selection


```{r}

economics_long %>% 
  filter(variable%in%c("pop","unemploy")) %>%
  ggplot(aes(x=date, y=value01)) + 
    geom_line(aes(linetype=variable))+
    scale_linetype_manual(values=c(2,3))

economics_long %>% 
  filter(variable%in%c("pop","unemploy")) %>%
  ggplot(aes(x=date, y=value01)) + 
    geom_line(aes(linetype=variable))+
    scale_linetype_manual(values=c("dotted","dashed"))


```




<br><br>

::: color-box 
::: 






# ✅ Best Practices 

## Color vs. shape


### Find the red dot

- based on <https://socviz.co/lookatdata.html> (Figure 1.18)

```{r echo=FALSE, message=FALSE, fig.cap="*Searching for the red dot becomes progressively harder.*"}

set.seed(2)
d20 <- tibble(x=sample(x = seq(0,30,by=0.01), size=20),
                  y=sample(x = seq(0,100,by=0.01), size=20),
                  color=c(rep("B",times=19),"A"),
                  shape=c("X",rep("Y",times=19))
                  )

d100 <- tibble(x=sample(x = seq(0,30,by=0.01), size=100),
                  y=sample(x = seq(0,100,by=0.01), size=100),
                  color1=c(rep("B",times=99),"A"),
                  shape1=c(rep("Y",times=99),"X"),
                  color2=c(rep("A",times=51),rep("B",times=49)),
                  shape2=c(rep("Y",times=50),rep("X",times=50)),
                  color3=c(rep("B",times=25)            ,rep("A",times=26)    ,rep("B",times=24),rep("A",times=25)),
                  shape3=c(rep(c("X","Y"),length.out=25),"X",rep("Y",times=25),rep(c("X","Y"),length.out=24),rep("Y",times=25))                  
                  )


p1 <- ggplot(d20,aes(x=x,y=y,color=color))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe, N=20")+
  xlim(0,30)+ylim(0,100)

p2 <- ggplot(d100,aes(x=x,y=y,color=color1))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe, N=100")+
  xlim(0,30)+ylim(0,100)

p3 <- ggplot(d20,aes(x=x,y=y,shape=shape))+
  geom_point(color="#F8766D")+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Form, N=20")+
  xlim(0,30)+ylim(0,100)

p4 <- ggplot(d100,aes(x=x,y=y,shape=shape1))+
  geom_point(color="#F8766D")+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Form, N=100")+
  xlim(0,30)+ylim(0,100)


p5 <- ggplot(d100,aes(x=x,y=y,color=color2,shape=shape2))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(legend.position="none",
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe+Form, N=100")+
  xlim(0,30)+ylim(0,100)


library(patchwork)

p1|p2|p3|p4|p5

```


### Find the red dot (2)


```{r echo=FALSE, message=FALSE}

p5 <- ggplot(d100,aes(x=x,y=y,color=color3,shape=shape3))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("Farbe+Form, N=100")+
  xlim(0,30)+ylim(0,100)

p6 <- ggplot(d100,aes(x=x,y=y,color=paste(color3,shape3)))+
  geom_point()+
  theme_minimal(base_size = 8)+
  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank())+
  ggtitle("3 Farben, N=100")+
  xlim(0,30)+ylim(0,100)

p5|p6
```




::: {.callout-tip icon=false}
## Color dominates for categorical variables  

see e.g.
[Channel Expressiveness Types and Effectiveness Ranks](https://www.cs.ubc.ca/~tmm/vadbook/eamonn-figs/fig5.1.pdf)


from: 
Tamara Munzner. Visualization Analysis and Design. A K Peters Visualization Series, CRC Press, 2014. <https://www.cs.ubc.ca/~tmm/vadbook/>


::: 


<br><br>

::: color-box 
::: 


## Design for colorblind readers

Roughly 8% of men and 0.5% of women have some form of color vision deficiency. Strategies:

* do use professionally composed color palettes (suited for color blindness)
  - such as, the **Okabe-Ito** palette or **viridis**, or **scico**-palettes.
- **Redundantly encode** with shape and/or linetype.
- Check your figure with `colorspace::cvd_emulator()` or <https://www.color-blindness.com/coblis-color-blindness-simulator/>.




<br><br>

::: color-box 
::: 




## Match palette type to data type

- **Categorical** → qualitative palettes (ColorBrewer Set2, Okabe-Ito). 
- **Ordered / continuous** → sequential palettes (e.g. viridis family) with perceptually uniform lightness steps.
- **Diverging** → diverging palettes (scico `berlin`, `vik`; ColorBrewer `RdBu`) when the midpoint has meaning (e.g., zero, an average).



- [Scientific colour maps by type ](https://s-ink.org/scientific-colour-maps)  <- VISIT   
- [Colour map guideline](https://s-ink.org/colour-map-guideline)  


<br><br>

::: color-box 
::: 


### `#endrainbow`

- **DO NOT USE** 'basic' **rainbow** color palettes 
  - Perceptual non-uniform
  - Artificial boundaries
  - Accessibility issues
  - Ambiguous ordering 
- <https://www.fabiocrameri.ch/endrainbow/>
- Crameri, F., Shephard, G.E. & Heron, P.J. The misuse of colour in science communication. Nat Commun 11, 5444 (2020). 
<https://doi.org/10.1038/s41467-020-19160-7>
- <https://clauswilke.com/dataviz/color-pitfalls.html>

<br><br>

::: color-box 
::: 


### Packages with color palettes

* `scico` -- <https://github.com/thomasp85/scico> uses color maps from <https://www.fabiocrameri.ch/colourmaps/>
* `cols4all` -- `cols4all::c4a_gui()`  -- <https://github.com/mtennekes/cols4all>
* `colorBlindness` <https://cran.r-project.org/web/packages/colorBlindness/vignettes/colorBlindness.html>
* ...

Extensive list: <https://github.com/EmilHvitfeldt/r-color-palettes>

<br><br>

::: color-box 
::: 


## Limit the number of categories

- **Color**: > 7–8 categories becomes very hard to distinguish.
- **Shape**: > 5–6 shapes is cognitively demanding.
- If you have many groups, consider faceting (`facet_wrap()`) instead (**later in the course**)


<br><br>
<br>

>**Categorical** color palettes in data visualization should not have more than **5 colors** (this is a fairly well-discussed "rule" of the field). In addition to this, **redundant encodings** should follow the same rules.<br>
> <https://observablehq.com/@frankelavsky/no-use-of-color-alone-in-data-visualization>


<br><br>

<br><br>

::: color-box 
::: 

## Maintain consistency across figures

Use the same color–category mapping throughout a report or presentation. Define your palette as a named vector once and reuse it:

```{r palette-reuse}
# named vector: c(name = value,...)
species_colors <- c(
  "setosa"     = "#D55E00",
  "versicolor" = "#0072B2",
  "virginica"  = "#009E73"
)
species_colors


ggplot(iris,aes(x=Petal.Width,y=Petal.Length,color=Species)) +
    geom_point()+
    scale_color_manual(values=species_colors)


# only versicolor - but same color assignment 
iris %>% 
  filter(Species=="versicolor") %>%
  ggplot(aes(x=Petal.Width,y=Petal.Length,color=Species)) +
      geom_point()+
      scale_color_manual(values=species_colors)



# Exmple: get colors from predefined palettes e.g. scico
# search for color palette access functions 
scico(n=5,palette="bilbao",begin=0,end=0.8)

my_cols <- c("cat1"="#4C0000" 
            ,"cat2"="#A46E59" 
            ,"cat3"="#C4BFAC"
)
```


<br><br>

::: color-box 
::: 


## Point shapes

- search for 'strong visual boundaries' (o,+)
- use (highlighting) colors

> "The hollow circle is a flexible and robust plotting symbol"
> Krzywinski, M., Wong, B. Plotting symbols. Nat Methods 10, 451 (2013). <https://doi.org/10.1038/nmeth.2490>



```{r echo=FALSE}
#from ggpubr::show_line_types()

lt <- c("blank", "solid", "dashed", "dotted", "dotdash",  "longdash", "twodash")

d <- data.frame(lt = factor(lt, levels = lt))

ggplot() + scale_x_continuous(name = "", limits = c(0, 1), 
    breaks = NULL) + scale_linetype_identity() + geom_segment(data = d, 
    mapping = aes(x = 0, xend = 1, y = lt, yend = lt, linetype = lt)) + 
    labs(title = "Line types available in R", y = "") + theme(axis.text.y = element_text(face = "bold", 
    color = "black"))+
    theme_minimal()+
    scale_y_discrete(labels=paste(0:6,lt,sep=' = '))

```



<br><br>

::: color-box 
::: 



<br><br><br><br>









# Bonus: Corporate Design - HEX RGB color definitions

## RGB or HEX

- RGB <https://en.wikipedia.org/wiki/RGB_color_model>
- HEX <https://www.color-hex.com>


```{r}

#create 3 colors by RGB definition
hgu_petrol      <- rgb(0, 108, 125, maxColorValue = 255)
hgu_lightgray   <- rgb(197, 198, 198, maxColorValue = 255)
hgu_lightgreen  <- rgb(174, 195, 11, maxColorValue = 255)


# Combine colors in a vector for plotting
hgu_3cols <- c(hgu_petrol,hgu_lightgray,hgu_lightgreen)


# Use colors in ggplot
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_color_manual(values=hgu_3cols)


# improved visibility of light colors with fill and shape 
ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
      geom_point(aes(fill=Species), shape=21)+
      scale_fill_manual(values=hgu_3cols)+
      theme_bw()


```

```{r}

# 3 HEX-Code Colors
cols <- c("#ff0000","#ffa500","#0000ff")

ggplot(data=iris, aes(x=Sepal.Width,y=Sepal.Length))+
  geom_point(aes(color=Species))+
  scale_color_manual(values=cols)

```

::: color-box 
::: 



<br><br><br><br>




# Additional resources

- <https://socviz.co/lookatdata.html>
- <https://en.wikipedia.org/wiki/Alpha_compositing>
- <https://en.wikipedia.org/wiki/Web_colors#Hex_triplet>
- <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html>
- <https://ggplot2.tidyverse.org/reference/aes_linetype_size_shape.html>
- <https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html>
- <https://lisacharlotterost.de/2016/04/22/Colors-for-DataVis/>
- <https://socviz.co/workgeoms.html>
- <https://ggplot2-book.org/scale-colour.html>
- <https://github.com/LKremer/ggpointdensity>
- Krzywinski, M., Wong, B. Plotting symbols. Nat Methods 10, 451 (2013). <https://doi.org/10.1038/nmeth.2490>
- Wong, B. Avoiding color. Nat Methods 8, 525 (2011). <https://doi.org/10.1038/nmeth.1642>
- Wong, B. Points of view: Points of review (part 2). Nat Methods 8, 189 (2011). <https://doi.org/10.1038/nmeth0311-189>
- <https://cran.r-project.org/web/packages/colorBlindness/vignettes/colorBlindness.html>
- <http://mkweb.bcgsc.ca/biovis2012/krzywinski-visualizing-biological-data.pdf>
- <https://blog.datawrapper.de/10-ways-to-use-fewer-colors-in-your-data-visualizations/>
- <https://observablehq.com/@frankelavsky/no-use-of-color-alone-in-data-visualization>
- Gehlenborg, N., Wong, B. Mapping quantitative data to color. Nat Methods 9, 769 (2012). <https://doi.org/10.1038/nmeth.2134>  
- Wong, B. Points of view: Points of review (part 2). Nat Methods 8, 189 (2011). <https://doi.org/10.1038/nmeth0311-189>  
- <https://colorusage.arc.nasa.gov/GuidelinesHierarchy.php>   
- <https://medialab.github.io/iwanthue/>     
- <https://clauswilke.com/dataviz/color-pitfalls.html>

#  Session Info

```{r}

sessionInfo()

```
