---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r format(Sys.time(), '%Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"    
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 006: Scientific figures

- Necessary features of a graph
- remove <a href=https://en.wikipedia.org/wiki/Chartjunk>'chartjunk'</a>
- consistency between plots


## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
  
pacman::p_load("data.table")
pacman::p_load("ggplot2")

```


# Features of a graph

- <a href=https://www.clips.edu.au/displaying-data>Common features of a graph</a> (Communication skills for science students, The University of Queensland)



# Axis

## Axis labels

```{r}
#simple
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    xlab("Petal width (cm)")+
    ylab("Petal length (cm)")


#general
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")


#uses ?plotmath [greek symbols]
ggplot(data = iris, aes(x = Petal.Width, y = Sepal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name=quote(alpha[width]~(m)))+
    scale_y_continuous(name=quote(beta[length]~(m)))

```

### `ggtext`

- https://github.com/wilkelab/ggtext

```{r}
pacman::p_load("ggtext")

ggplot(data = iris, aes(x = Petal.Width, y = Sepal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name="**Petal width (cm)**")+                                     # Markdown
    scale_y_continuous(name="<i style='color:orange'>***Petal Length<br>(cm)***</i> ")+  # HTML
      theme(
        axis.title.x = element_markdown(),
        axis.title.y = element_markdown(),
        plot.caption = element_markdown(),
      )    +
    labs(caption="Petal area (cm<sup>2</sup>) not shown.")                               # Markdown 


```


> Importantly, the gridtext package that provides the rendering support implements only an extremely **limited** subset of Markdown/HTML/CSS. It currently can make text bold or italics, can change the font, color, or size of a piece of text, can place text as sub- or superscript, and has extremely rudimentary image support. No other features are currently supported. As a general rule, any Markdown, HTML, or CSS feature that isn’t shown in any of the ggtext or gridtext documentation likely doesn’t exist.


## Axis limits + ticks

```{r}

#limits
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name="Petal width (cm)", limits=c(-1,5))+
    scale_y_continuous(name="Petal length (cm)", limits=c(-1,20))


# x-ticks
seq(-1,5,by=1.5)

# y-ticks
seq(-1,20,by=1.5)

# Breaks = ticks
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name="Petal width (cm)", limits=c(-1,5), breaks=seq(-1,5,by=1.5))+
    scale_y_continuous(name="Petal length (cm)", limits=c(-1,20), breaks=seq(-1,20,by=1.5))


```



## `coord_*`


```{r}
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name="Petal width (cm)", limits=c(-1,5), breaks=seq(-1,5,by=1.5))+
    scale_y_continuous(name="Petal length (cm)", limits=c(-1,20), breaks=seq(-1,20,by=1.5))+
    coord_fixed()
```


- `?coord_fixed` 
  - `coord_flip` - rotation (*flip*) of axis by 90°
  - `coord_polar` - transformation in polar coordinates
  - `coord_map` - for projection on a 2D world map
  - ...



# Legends

>  In ggplot2, you don’t directly control the legend; instead you set up the data so that there’s a clear mapping between data and aesthetics, and a legend is generated for you automatically. This can be frustrating when you first start using ggplot2, but once you get the hang of it, you’ll find that it saves you time, and there is little you cannot do. If you’re struggling to get the legend you want, it’s likely that your data is in the wrong form.
> <small> Q: https://ggplot2-book.org/scales.html </small>


```{r}

# Change legend title = name
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")+
    coord_fixed()+
    scale_colour_discrete(name="Iris species")    #Keeps colors as is, but rename title


# Change legend item names = labels

ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")+
    coord_fixed()+
    scale_colour_discrete(labels=c("1","2","3"))   #Keeps colors as is, but rename items


```


## Combine legends

```{r}

# Global color specification 
# Combines point and line legend

ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length, color = Species)) + 
    geom_point()+
    geom_smooth(method="lm")+
    scale_colour_discrete(name="Iris species") 

```


```{r}
#Shape and color are automatically combined in a single legend 
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species, shape=Species))

# unequal names and labels lead to separate legends
    # here: by manual renaming of color legend
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species, shape=Species))+
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")+
    coord_fixed() +
    scale_colour_discrete(name="Species ID", labels=c("1","2","3"))     

# rename both collapses legend to a single one
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species, shape=Species))+
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")+
    coord_fixed() +
    scale_colour_discrete(name="Species ID", labels=c("1","2","3"))    + 
    scale_shape_discrete(name="Species ID", labels=c("1","2","3"))   
```



```{r}
# add custom color and rename + change linetype and name
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species, shape=Species))+
    geom_smooth(method="lm",aes(linetype="Linear model"), color="black")+
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")+
    scale_colour_discrete(name="Species ID", labels=c("1","2","3"))    + 
    scale_shape_discrete(name="Species ID", labels=c("1","2","3"))     +
    scale_linetype_manual(name="Across species fit", values="dashed")

```




# Annotations / Direct labelling

- <a href=https://ggplot2-book.org/annotations.html#direct-labelling>'*direct labelling*'</a>.
- sometimes *the better legend*
- no need to 'search' for the legend box 
- `geom_text`, `geom_label` if you want to reuse data mappings
- `annotate` more flexible

```{r}
# add custom color and rename + change linetype and name
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species, shape=Species))+
    geom_smooth(method="lm", color="black")+
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")+
    scale_colour_discrete(name="Species ID", labels=c("1","2","3"))    + 
    scale_shape_discrete(name="Species ID", labels=c("1","2","3"))     +
      annotate(
        geom = "curve", x = 0.5, y = 4, xend = 0.75, yend = 3, 
        curvature = -.3, arrow = arrow(length = unit(2, "mm"))
      ) +
      annotate(geom = "text", x =0.45, y = 4, label = "Linear model", hjust = "right")

```


## Helper Packages

- not always too comfortable in base ggplot2
  - `directlabels` https://tdhock.github.io/directlabels/
  - `ggforce` (https://ggforce.data-imaginist.com/)
  - `ggrepel` (e.g. avoids overlapping) (https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html)
  - `gghighlight` https://github.com/yutannihilation/gghighlight
  - `geomtextpath` https://github.com/AllanCameron/geomtextpath


### `ggforce`


```{r}
#install.packages("ggforce")
library(ggforce)

# use annotation
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species, shape=Species),show.legend=FALSE)+
    geom_smooth(method="lm", color="black")+
    geom_mark_ellipse(aes(label = Species, group = Species, color=Species),show.legend=FALSE)+ #ggforce-Function
    scale_x_continuous(name="Petal width (cm)")+
    scale_y_continuous(name="Petal length (cm)")+
    annotate( #Draw curved arrow
        geom = "curve", x = 0.5, y = 4, xend = 0.75, yend = 3, 
        curvature = -.3, arrow = arrow(length = unit(2, "mm"))
        ) +
     annotate( #Print text at the end of arrow
        geom = "text", x =0.45, y = 4, 
        label = "Linear model\n across species", hjust = "right"
        )


```


### `directlabels`

- `directlabels` https://tdhock.github.io/directlabels/

```{r}
pacman::p_load(directlabels)


ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length, fill = Species)) + 
    geom_count(shape=21,alpha=0.75)+
    geom_smooth(method="lm", color="black")+        
    geom_dl(aes(label = Species, color = Species), method = "smart.grid")

```


### `geomtextpath`

- https://github.com/AllanCameron/geomtextpath


```{r}
pacman::p_load(geomtextpath)


ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_count(aes(fill = Species),shape=21,alpha=0.75)+
    geom_labelsmooth(method="lm",aes(label="Linear Model")
      ,color="black"
      ,size=3
      ,boxlinetype=0
      ,alpha=0.75
      )      
```






# Themes 

- custom design - 'cosmetics'

> Graphs have other features not strictly connected to the logical structure of the data being displayed. These include things like their background color, the typeface used for labels, or the placement of the legend on the graph. To adjust these, use the theme() function.
> <small> Q: https://socviz.co/workgeoms.html</small>






```{r}
# Save plot to a variable
p1 <- ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))

# plot by executing the variable  
p1 

# same as before with 'standard theme'
p1 + theme_gray()


# modifiy/extend existing plot with another theme
p1 + theme_minimal() #white bg

p1 + theme_bw() #white bg, black border

p1 + theme_light() #white bg, gray border

p1 + theme_classic() #white bg, no grid, no 'border' only axis


```



```{r}
# additional packages
#install.packages("ggthemes")
library(ggthemes)

p1 + theme_few() #larger font size, no grid, border

p1 + theme_hc() # only horizontal grid, font size modification, modification of legend placement


p1 + theme_gdocs() + # google docs theme
     scale_color_gdocs() # plus google color palette


# For the complete list of themes in the package check
ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]

```


- `ggthemr` https://github.com/cttobin/ggthemr 
- ... 



## Customize a theme (1)


```{r eval=FALSE}

theme_grey(
  base_size = 11,
  base_family = "",   #=sans; alternatives: Times, Courier,...
  base_line_size = base_size/22,
  base_rect_size = base_size/22
)

```



```{r}

# Increase base size (e.g. for a presentation)
p1 +
    theme_classic(base_size = 20)


# change font type to serif style and slightly decrease font size, e.g. for journal publications with default font size 10pt
p1 +
   theme_classic(base_size = 10,
                  base_family="Times")



```


## Customize a theme (2)

- There are many options ... https://ggplot2.tidyverse.org/reference/theme.html



### Legend: Position and border


```{r}
# Legend to bottom
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    theme(legend.position  = "bottom" ) 

# Legend at 15% X and 85% Y [default justification: center of legend]
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    theme(legend.position  = c(0.15,0.85) ) #percentage of figure area

# Add border to legend Background-Box
    # Change justification to 'left' - Box left Side
    ##Justification defines which side of the legend that the legend.position coordinates refer to
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    theme(
          legend.position  = c(0.15,0.85), 
          legend.justification = "left",                          
          legend.background = element_rect(color="black")    #Add border to legend Background-Box
          ) 

# Remove fill color of background box
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    theme(
          legend.position  = c(0.15,0.85),   
          legend.justification = "left",  
          legend.background = element_rect(color="black", fill=NA) #remove fill color (default: "white")
          ) 

# Change legend direction | move legend slightly more left
ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
    geom_point(aes(color = Species))+
    theme(
          legend.position  = c(0.05,0.85),                             
          legend.justification = "left",            
          legend.background = element_rect(color="black", fill=NA), 
          legend.direction = "horizontal"                            #change legend direction 
          ) 


```

- there is more to legends...

```{r eval=FALSE}

legend.background, 
legend.margin,
legend.spacing,
legend.spacing.x,
legend.spacing.y,
legend.key,
legend.key.size,
legend.key.height,
legend.key.width,
legend.text,
legend.text.align,
legend.title,
legend.title.align,
legend.position,
legend.direction,
legend.justification,
legend.box,
legend.box.just,
legend.box.margin,
legend.box.background,
legend.box.spacing,

```


## Your theme

- select a reference theme and modify 
- most ggplot2 themes are themselves only modifications of the standard theme: `theme_gray`,
dessen Definition unter: <a href=https://github.com/tidyverse/ggplot2/blob/master/R/theme-defaults.r>Link</a> 



```{r}

theme_DS <- function(base_size = 10,                           #set default values for the 4 "base parameters"
                      base_family = "Palatino",
                      base_line_size = base_size / 22,
                      base_rect_size = base_size / 22){
  theme_classic(base_size = base_size,                               # call your favorite theme and pass the input paremters
                base_family = base_family,
                base_line_size = base_line_size,
                base_rect_size = base_rect_size
                ) %+replace%                                     # add customized 'theme()' options (replace)
    theme(
      axis.title = element_text(                                 # modify axis title (name text)
        color = rgb(0, 108, 125, maxColorValue = 255),           # hgu_petrol      <- rgb(0, 108, 125, maxColorValue = 255)
        size = rel(1.0)),                                        # relative size to base size = 100% = 10pt
      axis.text = element_text(                                  # modify axis text (tick text)
        color = rgb(56, 56, 56, maxColorValue = 255),            # dark gray
        size = rel(0.8)),                                        # relative size to base size = 80% = 8pt
      panel.grid.major = element_line(                           # add major grid lines
        color = rgb(197, 198, 198, maxColorValue = 255),         # hgu_lightgray   <- rgb(197, 198, 198, maxColorValue = 255)
        linetype = "dotted",                                     # dotted line
        size = base_line_size/1.5),                              # slighlty decreased line thickness
      axis.line = element_line(                                  # modifiy axis lines
        color = rgb(197, 198, 198, maxColorValue = 255),         # hgu_lightgray   <- rgb(197, 198, 198, maxColorValue = 255)
        arrow=arrow(type = "closed",length = unit(0.125, "cm"))),# draw tiny arrow at the end of the axis lines      
      legend.position = "bottom",                                # move legend to bottom
      legend.key.size = unit(0.25,"cm"),                         # decrease legend key size to a square of 0.25 cm
      complete = TRUE                                            # define this as a 'complete theme' - in this case all objects not defined here were inherited from the "superior theme"
    )
}


```



```{r}

p1+theme_DS()

```



```{r eval=FALSE}
# set your theme as a default in the current R session.

theme_set(theme_DS())

```



# Best practices

- reduce 'chartjunk' to highlight data (https://en.wikipedia.org/wiki/Chartjunk)
- 'Simplify to clarify' (Wong, B. Simplify to clarify. Nat Methods 8, 611 (2011). https://doi.org/10.1038/nmeth.1660)
- 'effective' graphics (https://stat545.com/effective-graphs.html)

## Grid Lines

- get rid of unnecessary grid lines
- try it out and compare with your *message*


### Example (1)


```{r}
# library("rdwd")
# #Select data from the DWD CDC FTP Server
#   #current=TRUE - check current FTP Server Index to gather up-to-date file information
# link <- selectDWD("Geisenheim", res="monthly", var="kl", per="historic", current=TRUE)
# #Download data from the DWD CDC FTP Server
# file <- dataDWD(link, read=FALSE, dir="DWDdata", quiet=TRUE, force=NA)
# #Process data from the DWD CDC FTP Server to a data.frame
# clim <- readDWD(file, varnames=TRUE)

# #Change data.frame to data.table
# clim <- as.data.table(clim)

# fwrite(clim,file="clim_DWD_Geisenheim_historic_monthly.csv")


clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")
clim[,"MESS_DATUM":=as.POSIXct(MESS_DATUM)] 
clim[,"month":=month(MESS_DATUM)]

pB1 <- ggplot(clim[month=="1"][MESS_DATUM >= as.POSIXct("1980-01-01") & MESS_DATUM < as.POSIXct("2017-01-01")], 
        aes(x=MESS_DATUM,y=MO_TT.Lufttemperatur))+
    geom_line()+
    geom_smooth(method="lm")+
    scale_x_datetime(name="t (Month/Year)",date_breaks = "5 year", date_labels = "%m/%Y")+
    ylab("January Monthly-Mean Air Temperature (°C)")+
    theme_classic()

```

- Plot without grid - when only the relative trend is relevant for your message 


```{r}

pB1

```


- Plot with horizontal grid only - if the absolute values are also  of interest, but not the exact timing 

```{r}
pB1  + 
    theme(panel.grid.major.y=element_line()) 

```



### Example (2)

- Compare February and May
- (Figure 2): https://www.perceptualedge.com/articles/dmreview/grid_lines.pdf 



### Example (3) - Scatterplots

- full grid when you want to talk about a specific *block*
- trend is clearly visible without grid
- but if you want to know the specific *Petal width* that relates to *Petal length* between 4-5\,cm, the grid can help

```{r}

pB3 <- ggplot(iris, aes(x=Petal.Width, y=Petal.Length))+
        geom_point()+
        theme_classic()+
        scale_x_continuous(name="Petal width (cm)",breaks=seq(0,3,by=0.25), limits=c(0,2.75))+
        scale_y_continuous(name="Petal length (cm)",breaks=0:7, limits=c(0,7))
        


pB3

pB3  + 
    theme(panel.grid.major=element_line()) 



```


### Example (4)

- 'helper lines' instead of a full grid for orientation


```{r}
# add freezing point - if this is relevant to your study 
pB1+
    geom_hline(yintercept = 0, linetype="dashed", color="grey")


```


## Axis limits

- Include full range or not?
- in general: Do not *abuse* the various options to *mislead* the viewer
- bar plots and histograms should include the *zero*
  - horizontal lines can help pointing out smaller differences (see above) 
  - maybe a threshold or reference value besides *0* would also be a valid limit
- scatter plots and line plots are often used to show trends in or between data and not their absolute values, hence the zero-baseline is not always relevant


## *Colors with a meaning*

- People usually associate specific colors with *feelings* or *natural phenomena* 
- use this to your benefit - to improve the perception speed
- i.e. warm colors for summer temperature, shades of blue for data related to water, indicate higher values with higher color intensities
- if possible, do not reuse the same color for a different variable. Keep colors consistent within a document




## 'Ten guidelines for effective data visualization in scientific publications'

> Christa Kelleher, Thorsten Wagener,
> Ten guidelines for effective data visualization in scientific publications,
> Environmental Modelling & Software,
> Volume 26, Issue 6,
> 2011,
> Pages 822-827,
> ISSN 1364-8152,
> https://doi.org/10.1016/j.envsoft.2010.12.006
> (http://www.sciencedirect.com/science/article/pii/S1364815210003270)
> (https://scholar.google.de/scholar?cluster=14672598521261445236&hl=de&as_sdt=0,5&as_vis=1)


- ***Keep it simple*** 
- ***Visualize the purpose*** of the plot (details versus patterns) 
- ***Axis ranges***: absolute magnitude versus focus on differences 

```{r}


clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")
clim[,"MESS_DATUM":=as.POSIXct(MESS_DATUM)] 

clim[,"Year":=year(MESS_DATUM)]
clim_lt_sub <- clim[Year > 1951]
meanTDF <- clim_lt_sub[,.(meanT=mean(MO_TT.Lufttemperatur)),by=.(Year)]


pacman::p_load("patchwork") #will come to that later

p1 <- ggplot(meanTDF, aes(x=Year,y=meanT))+
  geom_col()+
  theme_minimal()+
  geom_smooth(method="lm")+
  ggtitle("Annual mean temperature increased")

p2 <- ggplot(meanTDF, aes(x=Year,y=meanT-meanT[1]))+
  geom_col()+
  theme_minimal()+
  ggtitle(paste("Mean T Difference to", min(meanTDF$Year)))+
  geom_smooth(method="lm")


p1|p2

```

- ***Data transformation***: plotting on log-scale to catch outliers / to capture data 
- ***Aspect ratio*** affects visual perception of rates of change 
- ***Handle overlapping points***: transparency / count plots 
- ***Missing data in line plots*** of time-series: do not connect lines with data-gaps 
- Use of ***summary statistic plots*** leads to simplicity; do not do boxplots for less than 5 data points  
- Axis ranges between plots: keep ***consistent*** axis ranges, if similar plots/data appear in a publication 
- ***Match color scheme and data*** 
  - sequential colors for quantitative data, with low values in lighter tints 
  - diverging colors for highlighting contrasts 
  - categories can use qualitative schemes 

# *"Message Trumps Beauty"*

> Rule 9: Message Trumps Beauty
> <small>Rougier NP, Droettboom M, Bourne PE (2014) Ten Simple Rules for Better Figures. PLOS Computational Biology 10(9): e1003833. https://doi.org/10.1371/journal.pcbi.1003833</small>






# Exercise 

```{r}
library("data.table")
library("ggplot2")

clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")
clim[,"MESS_DATUM":=as.POSIXct(MESS_DATUM)] 

clim[,"Year":=year(MESS_DATUM)]
clim_lt_sub <- clim[Year > 1951]
minmaxmeanTDF <- clim_lt_sub[,.(meanT=mean(MO_TT.Lufttemperatur),minT=min(MO_TT.Lufttemperatur),maxT=max(MO_TT.Lufttemperatur)),by=.(Year)]
m_minmaxmeanTDF <- melt(minmaxmeanTDF,id.vars="Year")

ggplot(m_minmaxmeanTDF, aes(x=Year,y=value, fill=variable))+
  geom_col(position="dodge")+
  geom_smooth(method="lm",aes(color=variable))+
  theme_minimal()


```

- Start from the above plot and try to improve it - show that the trend of min, max and mean air temperature increased since 1952  
- Try to annotate instead of using a legend (e.g. lines) (***Hint***: You could use the last year's data in a mapping by providing it as a new `data` input; `geom_text(data=m_minmaxmeanTDF[Year==max(Year)],...)`)
- Assume there is also a figure caption that contains some information, that might help you keeping the plot 'clean'/focused 

 
# Additional resources

- https://www.data-to-viz.com/
- https://scales.r-lib.org/
  - `pretty_breaks` -  https://scales.r-lib.org/reference/pretty_breaks.html
  - `extendet_breaks` - https://scales.r-lib.org/reference/extended_breaks.html
- https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html
- https://ggplot2.tidyverse.org/reference/theme.html
- https://www.data-imaginist.com/2019/the-ggforce-awakens-again/
- `ggfittext`https://github.com/wilkox/ggfittext


## **Nature Methods**: Points of View columns

- https://protocolsmethods.springernature.com/posts/43650-data-visualization-a-view-of-every-points-of-view-column
    - e.g. *Plotting Symbols*: https://www.nature.com/articles/nmeth.2490
    - https://www.nature.com/articles/nmeth.2337 (Figure3: Grid density)


#  Session Info

```{r}

sessionInfo()

```
