---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}

source("./src/custom.R")

```


# 007: Facets and Patchwork



# 🎯 Learning objectives

::: {.callout-note icon=false}
## Layered graphics

**... understanding concepts:**

* you understand how facets can support clarity (**small multiples**) or can be used in multidimensional visualization 


**... tool usage:**

* you can **grouping** a few `ggplot`s with the `patchwork` syntax and add labels to the subplots.  
* you can create **small multiples** with `facet_` functions   


:::


<br><br>

::: color-box 
::: 


# Prerequisites

```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
  
pacman::p_load("tidyverse")
pacman::p_load("patchwork")

```




<br><br>

::: color-box 
::: 


# Arranging different plots with `patchwork` 

> The goal of patchwork is to make it ridiculously simple to combine separate ggplots into the same graphic.
> <small><https://patchwork.data-imaginist.com/index.html></small>

## Generate 4 arbitrary plots

```{r}
# generate 4 arbitrary plots
# Boxplot
p1 <- ggplot(data = iris, aes(x = Species, y = Petal.Length)) + 
        geom_boxplot(aes(color = Species))+
        theme_minimal()+
        scale_y_continuous(name="Petal length (m)", limits=c(0,max(iris$Petal.Length)))+
        theme(legend.position="bottom")


# Scatterplot
p2 <- ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
        geom_point(aes(color = Species))+
        theme_minimal()+
        scale_x_continuous(name="Petal width (m)")+
        scale_y_continuous(name="Petal length (m)", limits=c(0,max(iris$Petal.Length)))+
        theme( legend.position="bottom")

# Density 1
p3 <- ggplot(iris, aes(Petal.Length, fill = Species)) + 
        geom_density(alpha = 0.5) +
        theme_minimal()+
        scale_x_continuous(name="Petal length (m)")+
        theme( legend.position="bottom")

# Density 2
p4 <- ggplot(iris, aes(Petal.Width, fill = Species)) + 
        geom_density(alpha = 0.5) +
        theme_minimal()+
        scale_x_continuous(name="Petal width (m)")+
        theme( legend.position="bottom")




```

## Syntax

- *adding* plots

```{r}

p1+p2

# automatic line breaks
p1+p2+p3+p4

``` 


<br><br>

::: color-box 
::: 

<br><br>

- Tabulating plots 

```{r}
#... in a row
p1|p2|p3|p4

#... in a column
p1/p2/p3/p4

``` 



<br><br>

::: color-box 
::: 

<br><br>

- brackets 

```{r}

(p1|p2|p3)/p4


``` 



<br><br>

::: color-box 
::: 

<br><br>

## `plot_layout`

- using `plot_layout` to 

```{r}

p1+p2+p3+p4 + plot_layout(nrow = 1, byrow = TRUE,)

``` 


<br><br>

::: color-box 
::: 

<br><br>

## `plot_annotation`

- using `plot_annotation` to assing labels to the subplots

```{r}

p1+p2+p3+p4 + plot_annotation(tag_levels = "A")

``` 



<br><br>

::: color-box 
::: 



## Merge legends 

- ... where possible
- identical legends are *collected*


```{r}

p1+p2+p3+p4 + 
  plot_layout(guides = 'collect')

```


<br><br>

::: color-box 
::: 



## Theme modifications - global

```{r}

p1+p2+p3+p4  + 
  plot_layout(guides = 'collect') & 
  theme(legend.direction="vertical")


```


<br><br>

::: color-box 
::: 



### Theme modifications - local

- use only one color legend in the collection 
- further legends do not provide additional information 

```{r}

p1+theme(legend.position = "none")+
p2+
p3+theme(legend.position = "none")+
p4+theme(legend.position = "none")+
  plot_layout(guides = 'collect') & 
  theme(legend.direction="vertical")

```


<br><br>

::: color-box 
::: 



### Modify grid

```{r}

(p1|p2)/(p3|p4)  / guide_area() +
  plot_layout(guides = 'collect', heights = c(2,2,1)) & 
  theme(legend.box = "horizontal")


```


<br><br>

::: color-box 
::: 



## *Advanced* `patchwork` 

For additional/more complex problems related to the arrangement of plots, please refer to the package vignettes  (<https://patchwork.data-imaginist.com/<) e.g.

- <https://patchwork.data-imaginist.com/articles/guides/assembly.html>
- <https://patchwork.data-imaginist.com/articles/guides/layout.html>


<br><br>

::: color-box 
::: 



## Alternative packages

If alternative packages and their tutorials are addressing your grid arrangement problem directly - use them instead.


### `cowplot`

- <https://wilkelab.org/cowplot/index.html>
- <https://wilkelab.org/cowplot/articles/index.html>

### `egg`

- <https://cran.r-project.org/web/packages/egg/vignettes/Overview.html>
- <https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html>



<br><br><br><br><br><br><br>


<br><br>

::: color-box 
::: 



# Facets for multidimensional visualizations in `ggplot2`

- no additional package necessary



## `facet_` - 1 dimension

- e.g. for comparing the 'same' data from different years, differnt regions, differnt ...


```{r}
clim <- read_csv("clim_DWD_Geisenheim_historic_monthly.csv",show_col_types = FALSE)
clim <- clim %>% 
          mutate("month"=month(MESS_DATUM)
                 ,"year"=year(MESS_DATUM)
                 )

# all in one plot - too crowded
clim %>% 
  filter(year > 2000 & year <= 2012) %>%
  ggplot(aes(x=month,y=MO_TT.Lufttemperatur))+
    geom_line(aes(color=factor(year)))

# 'wrap' onto a grid
clim %>% 
  filter(year > 2000 & year <= 2012) %>%
  ggplot(aes(x=month,y=MO_TT.Lufttemperatur))+
    geom_line(aes(color=factor(year)))+
    facet_wrap(year~.)


# place onto rows/cols of a 'grid'
clim %>% 
  filter(year > 2000 & year <= 2012) %>%
  ggplot(aes(x=month,y=MO_TT.Lufttemperatur))+
    geom_line(aes(color=factor(year)))+
    facet_grid(year~.)    #rows ~ cols [. = placeholder]

clim %>% 
  filter(year > 2000 & year <= 2012) %>%
  ggplot(aes(x=month,y=MO_TT.Lufttemperatur))+
    geom_line(aes(color=factor(year)))+
    facet_grid(.~year)   #rows ~ cols [. = placeholder]


```



<br><br>

::: color-box 
::: 

- Different data set `country`-wise data 


```{r}
# data from: https://r-graph-gallery.com/web-line-chart-small-multiple-all-group-greyed-out.html
pacman::p_load("tidyverse")
df1 <- read.csv("https://raw.githubusercontent.com/holtzy/R-graph-gallery/master/DATA/dataConsumerConfidence.csv") %>% 
  mutate(date=lubridate::my(Time)) %>% 
  select(-Time) %>% 
  pivot_longer(!date, names_to = "country", values_to = "value") %>% 
  na.omit()

df1

df1 %>% 
  ggplot(aes(x=date,y=value))+
    geom_line()+
    facet_wrap(.~country)   #rows ~ cols [. = placeholder]

```



<br><br>

::: color-box 
::: 


## Alternative application of facets: different data - same plot 

- requires data in *long format* as an input


```{r}

# covert to long format (i.e. tidyverse -> pivot_longer)
iris.m <- iris  %>% 
            pivot_longer(cols =  -c(Species))

iris.m

```



- To get an overview of the iris data set

```{r}
# free wrap
ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_wrap(name~.)

# put on a grid | use a column for each variable 
ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_grid(.~name)


# put on a grid | use a row for each variable
 # AND allow different y-axis [if justified]
ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_grid(name~., scales="free_y")


# wrap 
 # AND allow different y-axis [if justified]
ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_wrap(name~., scales="free_y")

```



<br><br>

::: color-box 
::: 



## `facet_` and 2 dimensions



```{r}
# ---------------------------------------------------------
# 1. DATA PREPARATION (LONG FORMAT)
# ---------------------------------------------------------
# Creating a dataset for 9 different weather stations
set.seed(123)

climate_wrap_data <- expand.grid(
  Month = 1:12,
  Year = as.factor(2021:2022),
  Station = paste("Station", LETTERS[1:9]) # 9 distinct locations
)

# Generate temperatures with different base levels and seasonal swings per station
station_id <- as.numeric(climate_wrap_data$Station)
climate_wrap_data$Temperature <- 35 + 
  (station_id * 3.5) +                                         # Base temp varies
  sin((climate_wrap_data$Month - 4) * pi / 6) * (10 + station_id) + # Seasonality varies
  rnorm(nrow(climate_wrap_data), mean = 0, sd = 2.5)           # Random noise


climate_wrap_data



# 1d wrap - split data by station
ggplot(climate_wrap_data
       , aes(x = Month, y = Temperature, color = Year)) +
  geom_line() +
  facet_wrap(Station~.,)


# 2d wrap - split by Station + Year (set 3 cols)
ggplot(climate_wrap_data
       , aes(x = Month, y = Temperature, color = Year)) +
  geom_line() +
  # The Small Multiples Wrap:
  # Takes the 9 stations + Years and wraps them into a 3x3 grid
  facet_wrap(Station~Year, ncol = 3)


# 2d grid - split by Station (rows) + Year (cols)
ggplot(climate_wrap_data
       , aes(x = Month, y = Temperature, color = Year)) +
  geom_line() +
  facet_grid(Station~Year)

```

<br><br>

::: color-box 
::: 


## More details on `facet_wrap` vs `facet_grid`  

- The differences between facet_wrap() and facet_grid():
  - <https://ggplot2-book.org/facet.html>


<br><br>

::: color-box 
::: 


## Some Options ...

- check with `?face_grid` or `?facet_wrap`

```{r}
facet_grid(
  rows = NULL,
  cols = NULL,
  scales = "fixed",
  space = "fixed",
  shrink = TRUE,
  labeller = "label_value",
  as.table = TRUE,
  switch = NULL,
  drop = TRUE,
  margins = FALSE,
  axes = "margins",
  axis.labels = "all",
)
```


<br><br>

::: color-box 
::: 

### `labeller` - facet-text options

- `?labellers`

```{r}


clim %>% 
  filter(year > 2000 & year <= 2012) %>%
  ggplot(aes(x=month,y=MO_TT.Lufttemperatur))+
    geom_line(aes(color=factor(year)))+
    facet_grid(.~year,labeller="label_both")   #adds colname to strip.text label

```


<br><br>

::: color-box 
::: 

### Thematic options

- `theme()` options (`?theme`), some examples:

```{r eval=FALSE}
...
  strip.background,
  strip.background.x,
  strip.background.y,
  strip.clip,
  strip.placement,
  strip.text,
  strip.text.x,
  strip.text.x.bottom,
  strip.text.x.top,
  strip.text.y,
  strip.text.y.left,
  strip.text.y.right,
  strip.switch.pad.grid,
  strip.switch.pad.wrap,
...

```


<br><br>

::: color-box 
::: 


::: {.callout-warning icon=false}
## Be careful with `scales="free_y"`

By setting `scales = "free_y"`, the Y-axis numbers will change for every panel to perfectly fit that station's data. 
This should only be considered if you are plotting different variables within each subplot (e.g. cf. iris long data). 
Otherwise, it eliminates between-plot comparability and can be visually deceptive. If the audience assumes all panels share the same baseline, they may misinterpret small swings in a nearly constant dataset as a massive spike - a trend that, in the broader context of the data, would actually be almost 'flat'.

:::


<br><br>

::: color-box 
::: 


# Exercise

::: {.callout-tip icon=false}
## 👉  patchwork and facetting

**Exercise (1)**

1. Use the following code to load the `tree`-dataset and transfer it into a `data.table` and to add two artificial groups (experiments):

```{r}
pacman::p_load("tidyverse")
pacman::p_load("patchwork")

str(trees)

trees <- as_tibble(trees) %>% 
            mutate("Group":=rep(c("1","2","3","4"), length.out=nrow(trees)))  #Assign 'artifical' Groups





```

2. Information about the data (<https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/trees.html>)   
3. Create a plot to each of the three variables (use different plot types) showing their distribution and merge theme to a single plot using `patchwork`  
4. Modify the axis titles to match scientific standards   
5. Optional: Customize the appearance according to your preferences    



**Exercise (2)**

0. Same data... 
1. Use `ggplot2` (facets) to create scatterplots of the `Volume~Height` split into subplots per `Group`.
  a. facet_grid colwise  
  b. facet_grid rowise  
  c. facet_wrap 3 cols
2. Modify the axis titles to match scientific standards   
3. Optional: Customize the appearance according to your preferences    

:::

<br><br>

::: color-box 
::: 

# Bonus: `gghighlight` 

- Option 1 with package `gghighlight`     
- Option 2: manually


```{r}
pacman::p_load("gghighlight")

# highlight the 'country' line but show all others too
ggplot(df1,aes(x=date, y=value, color=country)) +
  geom_line() +
  gghighlight(use_direct_label = FALSE,    # can also be used for direct labelling
              unhighlighted_params = list(colour = alpha("grey85", 1)) # configure how to show non highlighted data 
              ) +
  facet_wrap(.~country)


# manual realization in ggplot
df1_mod <- df1 %>% 
              mutate("country_facet"=country)

ggplot(df1_mod,aes(x=date, y=value, color=country)) +
  geom_line(data=df1,aes(group=country),color="grey") +    
  #background lines: with missing 'facet' variable in df1 all gray lines appear everywhere
  geom_line(aes(color=country))+  # colored lines from df1_mod with facet
  facet_wrap(.~country_facet)

```





<br><br>

::: color-box 
::: 


# Additional resources


- <https://ggplot2-book.org/facet.html>
- <https://ggplot2.tidyverse.org/reference/facet_grid.html>
- <https://ggplot2.tidyverse.org/reference/gg-add.html>
- <https://fishandwhistle.net/post/2018/modifying-facet-scales-in-ggplot2/>
- <http://www.sthda.com/english/wiki/ggplot2-violin-plot-quick-start-guide-r-software-and-data-visualization>
- <https://ggforce.data-imaginist.com/> [zoom to data]

```{r}

pacman::p_load("ggforce")
ggplot(iris, aes(x=Petal.Length,y=Petal.Width, colour = Species)) +
  geom_point() +
  facet_zoom(x = Species == "versicolor")

ggplot(iris, aes(x=Petal.Length,y=Petal.Width, colour = Species)) +
  geom_point() +
  facet_zoom(xlim = c(4, 6),
             zoom.size=1) 

ggplot(iris, aes(x=Petal.Length,y=Petal.Width, colour = Species)) +
  geom_point() +
  facet_zoom(ylim = c(1, 2),
             zoom.size=1) 


```




#  Session Info

```{r}

sessionInfo()

```