---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "2020"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 007: Facets and Patchwork

- grouping plots with `patchwork`
- facets for multidimensional visualization


## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

usePackage <- function(p)   
{ 
  if (!is.element(p, installed.packages()[,1])) 
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1]) 
  require(p, character.only = TRUE) 
} 
  
usePackage("data.table")
usePackage("ggplot2")
usePackage("patchwork")
usePackage("rdwd")

```


# Arranging different plots 


## `patchwork` 

> The goal of patchwork is to make it ridiculously simple to combine separate ggplots into the same graphic.
> <small>https://patchwork.data-imaginist.com/index.html</small>



```{r}
library("ggplot2")
library("patchwork")

# generate 4 arbitrary plots
# Boxplot
p1 <- ggplot(data = iris, aes(x = Species, y = Petal.Length)) + 
        geom_boxplot(aes(color = Species))+
        theme_minimal()+
        scale_y_continuous(name="Petal length (m)", limits=c(0,max(iris$Petal.Length)))+
        theme(legend.position="bottom")


# Scatterplot
p2 <- ggplot(data = iris, aes(x = Petal.Width, y = Petal.Length)) + 
        geom_point(aes(color = Species))+
        theme_minimal()+
        scale_x_continuous(name="Petal width (m)")+
        scale_y_continuous(name="Petal length (m)", limits=c(0,max(iris$Petal.Length)))+
        theme( legend.position="bottom")

# Density 1
p3 <- ggplot(iris, aes(Petal.Length, fill = Species)) + 
        geom_density(alpha = 0.5) +
        theme_minimal()+
        scale_x_continuous(name="Petal length (m)")+
        theme( legend.position="bottom")

# Density 2
p4 <- ggplot(iris, aes(Petal.Width, fill = Species)) + 
        geom_density(alpha = 0.5) +
        theme_minimal()+
        scale_x_continuous(name="Petal width (m)")+
        theme( legend.position="bottom")




```

## Syntax

- *adding* plots

```{r}

p1+p2

# automatic line breaks
p1+p2+p3+p4

``` 

- Tabulating plots 

```{r}
#... in a row
p1|p2|p3|p4

#... in a column
p1/p2/p3/p4

``` 

- brackets 

```{r}

(p1|p2|p3)/p4


``` 


## `plot_layout`

- using `plot_layout`

```{r}

p1+p2+p3+p4 + plot_layout(nrow = 1, byrow = TRUE)

``` 



### Merge legends 

- ... where possible
- identical legends are *collected*


```{r}

p1+p2+p3+p4 + 
  plot_layout(guides = 'collect')

```


### Theme modifications - global

```{r}
p1+p2+p3+p4  + 
  plot_layout(guides = 'collect') & 
  theme(legend.direction="vertical")


```

### Theme modifications - local

- use only one color legend in the collection 
- further legends do not provide additional information 

```{r}

p1+theme(legend.position = "none")+
p2+
p3+theme(legend.position = "none")+
p4+theme(legend.position = "none")+
  plot_layout(guides = 'collect') & 
  theme(legend.direction="vertical")

```

### Modify grid

```{r}

(p1|p2)/(p3|p4)  / guide_area() +
  plot_layout(guides = 'collect', heights = c(2,2,1)) & 
  theme(legend.box = "horizontal")


```



## *Advanced* `patchwork` 

For additional/more complex problems related to the arrangement of plots, please refer to the package vignettes  (https://patchwork.data-imaginist.com/) e.g.

- https://patchwork.data-imaginist.com/articles/guides/assembly.html
- https://patchwork.data-imaginist.com/articles/guides/layout.html

## Alternative packages

If alternative packages and their tutorials are addressing your grid arrangement problem directly - use them instead.


### `cowplot`

- https://wilkelab.org/cowplot/index.html
- https://wilkelab.org/cowplot/articles/index.html

### `egg`

- https://cran.r-project.org/web/packages/egg/vignettes/Overview.html
- https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html





# Facets for multidimensional visualizations in `ggplot2`

- no additional package necessary
- requires data in long format as an input


## `facet_grid` - 1 dimension



```{r}

library(data.table)
library(ggplot2)

iris <- as.data.table(iris)

# covert to long format 
iris.m <- melt(iris, id.vars="Species")

iris.m

```


- To get an overview of the iris data set

```{r}
# Approach 1: color boxplot by variable
ggplot(iris.m, aes(x=Species,y=value,color=variable))+
 geom_boxplot()

# Hard to compare across species

```

```{r}
# Approach 2: color boxplot by Species
ggplot(iris.m, aes(x=variable,y=value,color=Species))+
 geom_boxplot()

# not too bad here, but fails when variables are on more different scales

```

```{r}
# put on a grid | use a column for each variable 
ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_grid(.~variable)

#more or less similar to approach 2 
```


```{r}
# put on a grid | use a row for each variable
ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_grid(variable~.)


# put on a grid | use a row for each variable
 # AND allow different y-axis
ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_grid(variable~., scales="free_y")

```



- Alternative: `facet_wrap` for custom 'grid' (see below)




## `facet_grid` - 2 dimensions



```{r}
set.seed(1)
sinoid <- function(x) {rnorm(n=1,mean=sin(x*(2*pi/365)),sd=0.2)+5}
vsinoid <- Vectorize(sinoid)

dFACE <- data.table(
            Day = rep(c(1:365),times=6),
            Ring_Pair=factor(rep(c(1:3,1:3),each=365)),
            Treatment= rep(c("A","E"),each=365*3)     
          )


dFACE[,"Measurement":=vsinoid(Day)]


dFACE

```


```{r}
# Split by Ring_Pair [rows] + Treatment [cols]
  # plus: Color by Treatment
ggplot(dFACE, aes(x=Day, y=Measurement,color=Treatment))+
    geom_point()+
    facet_grid(Ring_Pair~Treatment)+
    scale_color_manual(values=c("blue","red"))+
    theme_classic() 


```


```{r}

ggplot(dFACE, aes(x=Day, y=Measurement,fill=Treatment))+
    geom_point(alpha=0.5,shape=21)+
    facet_grid(Ring_Pair~.)+
    scale_fill_manual(values=c("blue","red"))+
    theme_classic() 


```


## `facet_wrap`

- The differences between facet_wrap() and facet_grid():
  - https://ggplot2-book.org/facet.html



### `facet_wrap` - 1 dimension



```{r}

ggplot(dFACE, aes(x=Day, y=Measurement,fill=Ring_Pair))+
    geom_point(alpha=0.5,shape=21)+
    facet_wrap(.~Treatment)+
    scale_fill_viridis_d()+
    theme_classic() 
```

```{r}
# same plot 
ggplot(dFACE, aes(x=Day, y=Measurement,fill=Ring_Pair))+
    geom_point(alpha=0.5,shape=21)+
    facet_wrap(.~Treatment)+
    scale_fill_viridis_d()+
    theme_classic() 

```


### `facet_wrap` - 2 dimensions

```{r}

ggplot(dFACE, aes(x=Day, y=Measurement,color=Treatment))+
    geom_point()+
    facet_wrap(Ring_Pair~Treatment)+
    scale_color_manual(values=c("blue","red"))+
    theme_classic() 

```



```{r}

ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_wrap(variable~., scales="free")

```


## `facet_*` options

### `scales` - free axis scales

```{r}

ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_grid(variable~., scales="free_y")

```


```{r}

ggplot(iris.m, aes(x=Species,y=value, color=Species))+
 geom_boxplot()+
 facet_wrap(variable~., scales="free")

```


### `labels` - facet-text options

- `?labellers`

```{r}
ggplot(dFACE, aes(x=Day, y=Measurement,color=Treatment))+
    geom_point()+
    facet_grid(Ring_Pair~Treatment, label="label_both")+
    scale_color_manual(values=c("blue","red"))+
    theme_classic() 
```

For mathemaical symbols use, e.g.

> `label_parsed()` interprets the labels as plotmath expressions. 
> `label_bquote()` offers a more flexible way of constructing plotmath expressions. 




### Further options

- `?facet_wrap`: e.g. `strip.positon`
- `theme()` options (`?theme`), some examples:

```{r eval=FALSE}
  strip.background,
  strip.background.x,
  strip.background.y,
  strip.placement,
  strip.text,
  strip.text.x,
  strip.text.y,
  strip.switch.pad.grid,
  strip.switch.pad.wrap,

```



# Exercise (1)

1. Use the following code to load the 'tree'-dataset, transfer it into a `data.table` and to add two artificial groups (experiments):

```{r}
library("data.table")
library("ggplot2")
library("patchwork")

str(trees)

trees <- as.data.table(trees)

#Assig 'artifical' Groups
trees[,"Group":=rep(c("1","2"), length.out=nrow(trees))]


```

2. Information about the data (https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/trees.html)
3. Create a plot to each of the three variables showing their distribution and merge theme to a single plot using `patchwork`
4. Modify the axis titles to match scientific standards
5. Optional: Customize the appearance according to your preferences




# Exercise (2)


1. Load the historic climate data set from Geisenheim (s. code) and extract three years from this long-time data to a new data table (s. code example). You could use different years.
2. Transfer into a long-format data set for plotting.
3. Use `ggplot2` (facets) to create a plot illustrating two variables (`MO_TT.Lufttemperatur` und `MO_RR.Niederschlagshoehe`) by year.
4. Optional: Customize the appearance according to your preferences



```{r message=FALSE}
library("data.table")
library("ggplot2")
library("rdwd")

clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")


#Create a new year variable from date 
clim[,"Year":=year(MESS_DATUM)] # using year function (provided by data.table)
#clim[,"Year":=substr(MESS_DATUM,1:4)] #alternative extract first 4 digits from string
# More alternatives can be found with a quick google search: https://stackoverflow.com/questions/36568070/extract-year-from-date


# Extract 3 Years [2000-2002]
# Using Year
climSub <- clim[Year %in% c(2000:2002)]
#climSub <- clim[Year %in% c(1998,2001,2004)]

# ... should be 3x12=36
nrow(climSub)


# melt to long format using relevant columns only:
climSub.m <- melt(climSub[,c("MESS_DATUM","Year","MO_TT.Lufttemperatur","MO_RR.Niederschlagshoehe")], id.vars=c("MESS_DATUM","Year"))

str(climSub.m)

```






# Additional resources


- https://ggplot2-book.org/facet.html
- https://ggplot2.tidyverse.org/reference/facet_grid.html
- https://ggplot2.tidyverse.org/reference/gg-add.html
- https://fishandwhistle.net/post/2018/modifying-facet-scales-in-ggplot2/
- http://www.sthda.com/english/wiki/ggplot2-violin-plot-quick-start-guide-r-software-and-data-visualization
- https://ggforce.data-imaginist.com/ [zoom to data]

```{r}

library(ggforce)
ggplot(iris, aes(x=Petal.Length,y=Petal.Width, colour = Species)) +
  geom_point() +
  facet_zoom(x = Species == "versicolor")



ggplot(iris, aes(x=Petal.Length,y=Petal.Width, colour = Species)) +
  geom_point() +
  facet_zoom(xlim = c(4, 6),
             zoom.size=1) 



ggplot(iris, aes(x=Petal.Length,y=Petal.Width, colour = Species)) +
  geom_point() +
  facet_zoom(ylim = c(1, 2),
             zoom.size=1) 


```




#  Session Info

```{r}

sessionInfo()

```