---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
---
```{r echo=FALSE}

source("./src/custom.R")

```


# 008 Exporting Vector Graphics

## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
  
pacman::p_load("data.table")
pacman::p_load("ggplot2")


```

# Plot preparation


```{r message=FALSE}

clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")
clim[,"MESS_DATUM":=as.POSIXct(MESS_DATUM)] 
clim[,"month":=month(MESS_DATUM)]

pB1 <- ggplot(clim[month=="1"][MESS_DATUM >= as.POSIXct("1980-01-01") & MESS_DATUM < as.POSIXct("2017-01-01")], 
        aes(x=MESS_DATUM,y=MO_TT.Lufttemperatur))+
    geom_line()+
    geom_smooth(method="lm")+
    scale_x_datetime(name="t (Month/Year)",date_breaks = "5 year", date_labels = "%m/%Y")+
    ylab("January Monthly-Mean Air Temperature (°C)")+
    theme_classic()

```


# Export to file

- `ggsave()`
- if possible, export in the same dimensions in which you will use it (i.e. text/column width of a paper)


```{r}
# Options of the ggsave-function: ?ggsave
# ggsave(
#   filename,
#   plot = last_plot(),
#   device = NULL,
#   path = NULL,
#   scale = 1,
#   width = NA,
#   height = NA,
#   units = c("in", "cm", "mm"),
#   dpi = 300,
#   limitsize = TRUE,
# )

# Print the plot 
pB1

# save the last printed plot as 15x10cm PDF
ggsave("FigurePB1.pdf", width=15,height=10, unit="cm")


# save a plot (variable) without printing, as 15x10cm PNG with a 300 DPI resolution
ggsave("FigurePB1.png", plot=pB1, width=15,height=10, unit="cm", dpi=300)

ggsave("FigurePB1.svg", width=15,height=10, unit="cm")


```

- reccomended: **pdf**  -- *but* can be problematic in Microsoft Office
- common alternatives: ps, eps, **svg**

- Word and vector images <https://www.weston-tech.com/blog/vector-images-in-office-apps/>

> Device to use. Can either be a device function (e.g. png()), or one of "eps", "ps", "tex" (pictex), "pdf", "jpeg", "tiff", "png", "bmp", "svg" or "wmf" (windows only).
> <small><https://ggplot2.tidyverse.org/reference/ggsave.html></small>





# Exercise

- Create a plot of your choice and export it as a vector and as a raster graphics
- Compare the plot quality when zooming in 


# Additional resources

- <a href=https://youtu.be/-Fs2t6P5AjY>Vector versus raster graphics</a>

#  Session Info

```{r}

sessionInfo()

```
