---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r substr((Sys.Date()),1,4)`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 008 Exporting Vector Graphics


# Plot preparation


```{r message=FALSE}

usePackage <- function(p)   
{ 
  if (!is.element(p, installed.packages()[,1])) 
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1]) 
  require(p, character.only = TRUE) 
} 
  
usePackage("data.table")
usePackage("ggplot2")


clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")
clim[,"MESS_DATUM":=as.POSIXct(MESS_DATUM)] 


pB1 <- ggplot(clim[MESS_DATUM >= as.POSIXct("2015-01-01") & MESS_DATUM < as.POSIXct("2017-01-01")], 
        aes(x=MESS_DATUM,y=MO_TT.Lufttemperatur))+
        geom_line()+
        geom_smooth(method="lm")+
        scale_x_datetime(name="t (Month/Year)",date_breaks = "1 month", date_labels = "%m/%Y")+
        ylab("Monthly-Mean Temperature (°C)")+
        theme_classic()+
            theme(axis.text.x=element_text(angle = 45, vjust=0.5))

```


# Export to file

- `ggsave()`
- if possible, export in the same dimensions in which you will use it (i.e. text/column width of a paper)


```{r}
# Options of the ggsave-function: ?ggsave
# ggsave(
#   filename,
#   plot = last_plot(),
#   device = NULL,
#   path = NULL,
#   scale = 1,
#   width = NA,
#   height = NA,
#   units = c("in", "cm", "mm"),
#   dpi = 300,
#   limitsize = TRUE,
# )

# Print the plot 
pB1

# save the last printed plot as 15x10cm PDF
ggsave("FigurePB1.pdf", width=15,height=10, unit="cm")


# save a plot (variable) without printing, as 15x10cm PNG with a 300 DPI resolution
ggsave("FigurePB1.png", plot=pB1, width=15,height=10, unit="cm", dpi=300)



```

- most platform independent vector graphics format: pdf
- common alternatives: ps, eps

> Device to use. Can either be a device function (e.g. png()), or one of "eps", "ps", "tex" (pictex), "pdf", "jpeg", "tiff", "png", "bmp", "svg" or "wmf" (windows only).
> <small>https://ggplot2.tidyverse.org/reference/ggsave.html</small>





# Exercise

- Create a plot of your choice and export it as a vector and as a raster graphics
- Compare the plot quality when zooming in 





# Additional resources

- <a href=https://youtu.be/-Fs2t6P5AjY>Vector versus raster graphics</a>

#  Session Info

```{r}

sessionInfo()

```
