---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
   df-print: paged
editor_options: 
  chunk_output_type: console
---
```{r echo=FALSE}

source("./src/custom.R")

```


# 008 Exporting Vector Graphics



# 🎯 Learning objectives

::: {.callout-note icon=false}
## Layered graphics

**... understanding concepts:**

* you understand the difference between *vector* and *raster* images  


**... tool usage:**

* you can **save** a ggplot in a *vector* and *raster* images format with specified dimensions   


:::


<br><br>

::: color-box 
::: 

# Prerequisites

```{r message=FALSE}

if (!require("pacman")) install.packages("pacman")
  
pacman::p_load("ggplot2")
pacman::p_load("svglite")

```


<br><br>


::: color-box 
::: 

# Plot preparation


```{r message=FALSE}

p <- ggplot(iris,aes(x=Species,y=Petal.Length,color=Species))+
        geom_violin(alpha=0.2)+                         # (optional) supports sina-shape
        ggforce::geom_sina(jitter_y=FALSE,size=0.5)+    # deactivate value-jitter
        geom_boxplot(width=0.1          # adjust to place inside the violin
                    ,alpha=0.5          # adjust to see the raw data 
                    ,outliers = FALSE   # no 'duplicate' data points
                    ,color="gray30"
                    )+
        theme_minimal()+
        theme(legend.position = "none") +
        scale_color_viridis_d(begin=0.1,end=0.8)


```

<br><br>


::: color-box 
::: 

# Export to file

- `ggsave()`
- if possible, export in the same dimensions in which you will use it (i.e. text/column width of a paper)


```{r}
# Options of the ggsave-function: ?ggsave
# ggsave(
#   filename,
#   plot = last_plot(),
#   device = NULL,
#   path = NULL,
#   scale = 1,
#   width = NA,
#   height = NA,
#   units = c("in", "cm", "mm"),
#   dpi = 300,
#   limitsize = TRUE,
# )

# Print the plot 
p

# save the last printed plot as 9x8cm PDF
# if not 'plot' parameter is set.
ggsave("FigurePB1.pdf", width=9,height=8, unit="cm")

# save a plot (variable) without printing, as 9x8cm PNG with a 300 DPI resolution
# plot = p   
ggsave("FigurePB1.png", plot=p, width=9,height=8, unit="cm", dpi=300)

pacman::p_load("svglite")
ggsave("FigurePB1.svg", plot=p,width=9,height=8, unit="cm")


```

- reccomended: **pdf**  -- *but* can be problematic in Microsoft Office
- common alternatives: ps, eps, **svg**

- Word and vector images <https://www.weston-tech.com/blog/vector-images-in-office-apps/>

> Device to use. Can either be a device function (e.g. png()), or one of "eps", "ps", "tex" (pictex), "pdf", "jpeg", "tiff", "png", "bmp", "svg" or "wmf" (windows only).
> <small><https://ggplot2.tidyverse.org/reference/ggsave.html></small>



<br><br>


::: color-box 
::: 


# Exercise

::: {.callout-tip icon=false}
## 👉  Compare vector and raster graphics by zooming in


- Create a plot of your choice and export it as a vector and as a raster graphics
- Compare the plot quality when zooming in 
:::

<br><br>


::: color-box 
::: 

# Additional resources

- <a href=https://youtu.be/-Fs2t6P5AjY>Vector versus raster graphics</a>
- [cairo_pdf device for special character usage etc..](https://cran.r-project.org/web/packages/Cairo/)

#  Session Info

```{r}

sessionInfo()

```
