---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r Sys.Date()`"
date-format: "YYYY"
toc: true
toc-depth: 3
number-sections: false
highlight-style: tango
format: 
  html:
   embed-resources: true
   css: "./css/clean_qmd.css"
---
```{r echo=FALSE}

source("./src/custom.R")

```


# 009a Interactive Plots


## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}


if (!require("pacman")) install.packages("pacman")

pacman::p_load("data.table")  
pacman::p_load("ggplot2")  
pacman::p_load("plotly")
pacman::p_load("gganimate")
pacman::p_load("transformr")
# pacman::p_load("sf")


```


# A brief insight into interactive plots

- most popular interactive plotting library <a href=https://plotly.com/>`plotly`</a> (cross platform availability)


## `plotly`

- for R: <https://plotly.com/r/>
- no need to learn `plotly` syntax for simple graphs - convert ggplot to plotly: `ggplotly()`


```{r}

iris <- as.data.table(iris)

pScat  <- ggplot(iris,aes(x=Petal.Width,y=Petal.Length, color=Species))+
            geom_point()


ggplotly(pScat)


```

- "Mouse-Overs"
- Zoom
- png-export


### ***Limitations***
e.g. legend positioning

```{r}

pScat_legB <- pScat  +  
                theme(legend.position="bottom")

ggplotly(pScat_legB)


```


and `facet_`s (z.B. text outside of plot area). `plotly` has no equivalent functions - uses so called *subplots*


```{r}

# to long format
miris <- melt(iris, id.vars="Species")

# faceted plot
pF <- ggplot(miris,aes(x=variable,y=value, fill=variable))+
            geom_boxplot()+
            facet_wrap(.~Species)


ggplotly(pF)

```







## Animated ggplots -- `gganimate`


- *Storytelling* within a plot
- use with care, don't overdo it - only if your message benefits from the animation
- Additional layers for controlling the animation: *transition* und *ease_aes*
- run/global settings with `animate()`

A not too useful example:

```{r message=FALSE}


clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")
clim[,"MESS_DATUM":=as.POSIXct(MESS_DATUM)]

clim[,"Year":=year(MESS_DATUM)] 
clim[,"Month":=month(MESS_DATUM)]

pAnim <- ggplot(clim[Year > 1980], aes(x=Month, y=MO_TT.Lufttemperatur)) +
  geom_line(aes(color=MO_TT.Lufttemperatur))+
  geom_point(alpha = 0.7, show.legend = FALSE, aes(color=MO_TT.Lufttemperatur),size=5) +
  scale_colour_viridis_c(option="A") +
  scale_x_continuous(breaks=1:12)+
  theme_minimal()+
  # gganimate specific from here
  labs(title = 'Year: {frame_time}', x = 'Month', y = 'T (°C)') +
  transition_time(Year) +
  ease_aes('linear')

nYears <- length(unique(clim[Year > 1980]$Year))
nYears

# creat animation
animate(pAnim, nframes = nYears, fps=1)



```


More examples can be found e.g. at <https://gganimate.com/>.


# Additional resources

- <https://www.r-bloggers.com/comparing-plotly-ggplotly-plot-generation-times/>
- <https://www.musgraveanalytics.com/blog/2018/8/24/how-to-make-ggplot2-charts-interactive-with-plotly>
- <https://gganimate.com/articles/gganimate.html>
- <https://ggvis.rstudio.com/>

#  Session Info

```{r}

sessionInfo()

```