---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "2020"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "../custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 009a Interactive Plots


## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

usePackage <- function(p)   
{ 
  if (!is.element(p, installed.packages()[,1])) 
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1]) 
  require(p, character.only = TRUE) 
} 
  
usePackage("plotly")
usePackage("gganimate")
usePackage("transformr")
usePackage("sf")


```


# A brief insight into interactive plots

- most popular interactive plotting library <a href=https://plotly.com/>`plotly`</a> (cross platform availability)
- alternative R-packages, e.g. `ggvis`  (https://ggvis.rstudio.com/).


## `plotly`

- for R: https://plotly.com/r/
- no need to learn `plotly` syntax for simple graphs - convert ggplot to plotly: `ggplotly()`


```{r}

library("data.table")
library("ggplot2")
library("plotly")

iris <- as.data.table(iris)

pScat  <- ggplot(iris,aes(x=Petal.Width,y=Petal.Length, color=Species))+
            geom_point()


ggplotly(pScat)


```

- "Mouse-Overs"
- Zoom
- png-export


### ***Limitations***
e.g. legend positioning

```{r}

pScat_legB <- pScat  +  
                theme(legend.position="bottom")

ggplotly(pScat_legB)


```


and `facet_`s (z.B. text outside of plot area). `plotly` has no equivalent functions - uses so called *subplots*


```{r}

# to long format
miris <- melt(iris, id.vars="Species")

# facetted plot
pF <- ggplot(miris,aes(x=variable,y=value, fill=variable))+
            geom_boxplot()+
            facet_wrap(.~Species)


ggplotly(pF)

```







## Animated ggplots -- `gganimate`


- *Storytelling* within a plot
- use with care, don't overdo it - only if your message benefits from the animation
- Additional layers for controlling the animation: *transition* und *ease_aes*
- run/global settings with `animate()`

A not too useful example:

```{r message=FALSE}
library("data.table")
library("ggplot2")
library("gganimate")
library("transformr")
library("sf")


clim <- fread("clim_DWD_Geisenheim_historic_monthly.csv")

clim[,"Year":=year(MESS_DATUM)] 
clim[,"Month":=month(MESS_DATUM)]

pAnim <- ggplot(clim[Year > 1980], aes(x=Month, y=MO_TT.Lufttemperatur)) +
  geom_line(aes(color=MO_TT.Lufttemperatur))+
  geom_point(alpha = 0.7, show.legend = FALSE, aes(color=MO_TT.Lufttemperatur),size=5) +
  scale_colour_viridis_c(option="A") +
  scale_x_continuous(breaks=1:12)+
  theme_minimal()+
  # gganimate specific from here
  labs(title = 'Year: {frame_time}', x = 'Month', y = 'T (°C)') +
  transition_time(Year) +
  ease_aes('linear')

nYears <- length(unique(clim[Year > 1980]$Year))
nYears

# creat animation
animate(pAnim, nframes = nYears, fps=1)



```


More examples can be found e.g. at https://gganimate.com/.


# Additional resources

- https://www.r-bloggers.com/comparing-plotly-ggplotly-plot-generation-times/
- https://www.musgraveanalytics.com/blog/2018/8/24/how-to-make-ggplot2-charts-interactive-with-plotly
- https://gganimate.com/articles/gganimate.html
- https://ggvis.rstudio.com/

#  Session Info

```{r}

sessionInfo()

```