---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r substr((Sys.Date()),1,4)`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 009b Data management


## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

usePackage <- function(p)   
{ 
  if (!is.element(p, installed.packages()[,1])) 
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1]) 
  require(p, character.only = TRUE) 
} 
  
usePackage("ggplot2")
usePackage("data.table")
usePackage("scales")   #viridis_pal


```



## Load data and subset

```{r message=FALSE}
# library("rdwd")
# # 'recent' daily data from Geisenheim
# link <- selectDWD("Geisenheim", res="daily", var="kl", per="historical")
# file <- dataDWD(link, read=FALSE, dir="DWDdata", quiet=TRUE, force=NA)
# clim_lt <- readDWD(file, varnames=TRUE)
# clim_lt <- as.data.table(clim_lt)
# str(clim_lt)

#fwrite(clim_lt,"clim_DWD_Geisenheim_historic_daily.csv")

clim_lt <- fread("clim_DWD_Geisenheim_historic_daily.csv")


# Subset Dataset to relevant variables and slightly reformat (optional)
clim_lt_sub <- clim_lt[,c("MESS_DATUM","TMK.Lufttemperatur","UPM.Relative_Feuchte","FM.Windgeschwindigkeit","NM.Bedeckungsgrad","RSK.Niederschlagshoehe")]
colnames(clim_lt_sub) <- c("Date","T","phi","wind","clouds","rain")
clim_lt_sub[,"Date":=as.Date(Date)]
summary(clim_lt_sub)


```


## Introducing helper variables (faceting, aggregating, statistics)

-- example: year, month, day

```{r}

clim_lt_sub[,"Year":=year(Date)]
clim_lt_sub[,"Month":=month(Date)]
clim_lt_sub[,"Day":=mday(Date)]

#Alternatives 
#clim_lt_sub[,"Year":=substr(Date,1,4)]
# clim_lt_sub[,"Year":=as.numeric(format(Date,'%Y'))]
# clim_lt_sub[,"Month":=as.numeric(format(Date,'%m'))]
# clim_lt_sub[,"Day":=as.numeric(format(Date,'%d'))]

#library(stringi)
#with(clim_lt_sub, stri_split_regex(Date,pattern="-",simplify=TRUE)




# Subset Years
clim_lt_sub <- clim_lt_sub[Year > 1950]

```


## Introducing helper data-sets (aggregation)

```{r}

summary(clim_lt_sub)
#clim_lt_sub_YMEAN <- as.data.table(with(clim_lt_sub, aggregate(list(T_Air=T_Air, phi=phi), by=list(Year=Year), FUN=function(x) {mean(x, na.rm=TRUE)})))

clim_lt_sub_YMEAN <- clim_lt_sub[,lapply(.SD, function(x) {mean(x, na.rm=TRUE)}), by='Year', .SDcols=c('T','phi','wind','clouds','rain')]
clim_lt_sub_YSD   <- clim_lt_sub[,lapply(.SD, function(x) {sd(x, na.rm=TRUE)}),   by='Year', .SDcols=c('T','phi','wind','clouds','rain')]

clim_lt_sub_YMMEAN <- clim_lt_sub[,lapply(.SD, function(x) {mean(x, na.rm=TRUE)}), by=c('Year','Month'), .SDcols=c('T','phi','wind','clouds','rain')]
clim_lt_sub_YMSD   <- clim_lt_sub[,lapply(.SD, function(x) {sd(x, na.rm=TRUE)}),   by=c('Year','Month'), .SDcols=c('T','phi','wind','clouds','rain')]



# mean_sd <- function(x){
#                         list(mean = mean(x, na.rm=TRUE), sd = sd(x, na.rm=TRUE))
#                       }
# clim_lt_sub[, setNames( sapply(.SD, mean_sd), paste0(c("mean.", "sd."),rep(c('T','phi','wind','clouds','rain'),each=2))), by='Year', .SDcols=c('T','phi','wind','clouds','rain')]


```

## Creat Longformat Datasets for ggplot

```{r}

clim_lt_sub_melt <- na.omit(melt(clim_lt_sub, id.vars=c("Date","Year","Month","Day")))

#helper function
melt_and_narm_similar <- function(dataset) { na.omit(melt(dataset, id.vars=colnames(dataset)[colnames(dataset)%in%c("Date","Year","Month","Day")])) }


clim_lt_sub_YMEAN_melt  <- melt_and_narm_similar(clim_lt_sub_YMEAN)
clim_lt_sub_YSD_melt    <- melt_and_narm_similar(clim_lt_sub_YSD)
clim_lt_sub_YMMEAN_melt <- melt_and_narm_similar(clim_lt_sub_YMMEAN)
clim_lt_sub_YMSD_melt   <- melt_and_narm_similar(clim_lt_sub_YMSD)

#ADD SD to MEAN DF (merge)
clim_lt_sub_YMEAN_melt[,"sd":=clim_lt_sub_YSD_melt$value]
clim_lt_sub_YMMEAN_melt[,"sd":=clim_lt_sub_YMSD_melt$value]







```

## Some plotting options

```{r}

# Alltime data overview
ggplot(clim_lt_sub_melt, aes(x=variable,y=value))+
  geom_boxplot()+
  #geom_violin(draw_quantiles=c(0.25,0.5,0.75))+
  facet_wrap(~variable, scales="free")

# Alltime data overview - per Year
ggplot(clim_lt_sub_melt, aes(x=Year,y=value))+
  geom_boxplot(aes(group=Year))+
  #geom_violin(draw_quantiles=c(0.25,0.5,0.75))+
  facet_wrap(~variable, scales="free")+
  geom_smooth()+
  geom_vline(xintercept=2001)   #change of estimation of daily data


# YearMonth-Wise Mean data overview 
  # Outliers "removed"
ggplot(clim_lt_sub_YMMEAN_melt, aes(x=Year,y=value))+
  geom_boxplot(aes(group=Year))+
  #geom_point()+
  #geom_violin(draw_quantiles=c(0.25,0.5,0.75))+
  facet_wrap(~variable, scales="free")+
  geom_smooth()+
  geom_vline(xintercept=2001)

# Year-Wise Mean data overview 
ggplot(clim_lt_sub_YMEAN_melt, aes(x=Year,y=value))+
  geom_point()+
  facet_wrap(~variable, scales="free")+
  geom_smooth()+
  geom_vline(xintercept=2001)


# Year-Wise Mean data overview 
  #pointrange (sd) for RAIN yields unrealitic values)
ggplot(clim_lt_sub_YMEAN_melt, aes(x=Year,y=value))+
  geom_pointrange(aes(ymin=value-sd, ymax=value+sd))+       #pointrange - mean +/- sd 
  facet_wrap(~variable, scales="free")+
  geom_smooth()+
  geom_vline(xintercept=2001)


# Year-Wise Mean data overview with linear model
ggplot(clim_lt_sub_YMEAN_melt, aes(x=Year,y=value))+
  geom_point()+
  facet_wrap(~variable, scales="free")+
  geom_smooth(method="lm")+                               #linear model
  geom_vline(xintercept=2001)


```


### Towards a publication ready figure

- if you want to show variability between and within years and the overall trend
- box plots based on monthly averages

```{r}

unique(clim_lt_sub_YMMEAN_melt$variable)

levels(clim_lt_sub_YMMEAN_melt$variable)

#rename variables for plotting
clim_lt_sub_YMMEAN_melt$variable <- factor(clim_lt_sub_YMMEAN_melt$variable, 
                                            labels = c("T[air]~('°'*C)", 
                                                                        "phi~('%')",
                                                                        "windspeed~('m/s')",
                                                                        "coverage~('1/8')",
                                                                        "precipitation~('mm')"
                                                                        )
                                            )


seq_width <- 10
minY <- round(min(clim_lt_sub_YMMEAN_melt$Year)+seq_width/2,-1)
maxY <- round(max(clim_lt_sub_YMMEAN_melt$Year)-seq_width/2,-1)

#breaks includes start/end + decades
year_breaks <- c(min(clim_lt_sub_YMMEAN_melt$Year),
                seq(minY,maxY,by=seq_width),
                max(clim_lt_sub_YMMEAN_melt$Year)
                )


p <- 
ggplot(clim_lt_sub_YMMEAN_melt, aes(x=Year,y=value))+
    geom_boxplot(aes(group=Year),lwd=0.1,outlier.size = 0.1)+                     # modfiy for readability
    facet_wrap(~variable, scales="free", strip.position="left", ncol=1, 
                    labeller=label_parsed)+
    geom_smooth(method="lm", aes(linetype="Trendline", color="Trendline"), se=FALSE)+         #remove error | invisible and problematic for legends
    geom_vline(aes(xintercept=2001,linetype="DWD-Change", color="DWD-Change"), alpha=0.5)+
    scale_linetype_manual(name="",values=c(2,1))+               #ONLY necessary if "DWD Change" is used
    scale_colour_manual(name="", values=c("black","red"))+      #ONLY necessary if "DWD Change" is used
    scale_x_continuous(breaks=year_breaks, expand=c(0,0))   +         #Define x-axis ticks
    theme_minimal(base_size = 10)+
    theme(
              axis.title.y =element_blank()                             #remove "value" as y-axis label
              ,strip.placement = "outside"                              #use strip text as y-axis 
               ,legend.position="bottom"                               #move legend
             # ,legend.position=c(0.75,0.2)                               #move legend | if multi-column setup
              ,legend.key = element_rect(fill = 'white', color="white", size = 0.0)    #fill background and remove 
              ,legend.key.size = unit(0.5,"cm")    #fill background and remove border of "key"
             # ,legend.background = element_rect(color = 'black')        #draw box around legend
              ,legend.title = element_blank()                           #remove empty title entirely
              ,plot.margin = margin(t=0, r=0.35, b=0, l=0, "cm")                   #reduce plot margin ()
              ,legend.margin = margin(t=0, r=0, b=0, l=0, "cm")  
            )

# if provided by a publisher
single_column_width <- 8.5 #cm
max_column_height <- 29.7 - 2*3.75 #cm

ggsave(p,filename="Figure1.pdf", width=single_column_width, height=max_column_height, units="cm")



```


#### Plot 

```{r thumbnails=FALSE}

knitr::include_graphics("./Figure1.pdf")

```



#### Add Quarters of Year



```{r}

# quaŕters = cut month into 4 intervalls
#clim_lt_sub[,"Quarter":=paste0("Q",as.numeric(factor(cut(Month,4))))]
clim_lt_sub[,"Quarter":=paste0("Q",quarter(Date))]


# Calculate Means
clim_lt_sub_QMEAN <- clim_lt_sub[,lapply(.SD, function(x) {mean(x, na.rm=TRUE)}), by=c('Year','Quarter'), .SDcols=c('T','phi','wind','clouds','rain')]

#remove NAs and creat long format
clim_lt_sub_QMEAN_melt <- na.omit(melt(clim_lt_sub_QMEAN, id.vars=c("Year","Quarter")))


levels(clim_lt_sub_QMEAN_melt$variable)

#rename variables for facet strip text
clim_lt_sub_QMEAN_melt$variable <- factor(clim_lt_sub_QMEAN_melt$variable, 
                                            labels = c("T[air]~('°'*C)", 
                                                                        "phi~('%')",
                                                                        "windspeed~('m/s')",
                                                                        "coverage~('1/8')",
                                                                        "precipitation~('mm')"
                                                                        )
                                            )



p <- 
ggplot(clim_lt_sub_QMEAN_melt, aes(x=Year,y=value))+
    geom_point(aes(fill=Quarter,shape=Quarter), alpha=0.5)+                     # modfiy for readability
    facet_wrap(~variable, scales="free", strip.position="left", ncol=1, 
                    labeller=label_parsed)+
    geom_smooth(method="lm", aes(color=Quarter), se=FALSE)+         #remove error | invisible and problematic for legends (gray)
    scale_x_continuous(breaks=year_breaks, expand=c(0,0))   +         #Define x-axis ticks
    theme_minimal(base_size = 10)+
    theme(
              axis.title.y =element_blank()                             #remove "value" as y-axis label
              ,strip.placement = "outside"                              #use strip text as y-axis 
               ,legend.position="bottom"                               #move legend
             # ,legend.position=c(0.75,0.2)                               #move legend | if multi-column setup
              ,legend.key = element_rect(fill = 'white', color="white", size = 0.0)    #fill background and remove 
              ,legend.key.size = unit(0.5,"cm")    #fill background and remove border of "key"
             # ,legend.background = element_rect(color = 'black')        #draw box around legend
              ,legend.title = element_blank()                           #remove empty title entirely
              ,plot.margin = margin(t=0, r=0.35, b=0, l=0, "cm")                   #reduce plot margin ()
              ,legend.margin = margin(t=0, r=0, b=0, l=0, "cm")  
            )+
     scale_fill_manual(values=viridis_pal(option="A")(5)[c(1,3,4,2)])+
     scale_color_manual(values=viridis_pal(option="A")(5)[c(1,3,4,2)]) +  
    scale_shape_manual(values=c(21:24))

ggsave(p,filename="Figure1_1.pdf", width=8.5, height=max_column_height, units="cm")


#Line instead of points
p <- ggplot(clim_lt_sub_QMEAN_melt, aes(x=Year,y=value))+
    geom_line(aes(color=Quarter), alpha=0.5)+                     # modfiy for readability
    facet_wrap(~variable, scales="free", strip.position="left", ncol=1, 
                    labeller=label_parsed)+
    geom_smooth(method="lm", aes(color=Quarter),linetype="dashed", se=FALSE)+         #remove error | invisible and problematic for legends
    scale_x_continuous(breaks=year_breaks, expand=c(0,0))   +         #Define x-axis ticks
    theme_minimal(base_size = 10)+
    theme(
              axis.title.y =element_blank()                             #remove "value" as y-axis label
              ,strip.placement = "outside"                              #use strip text as y-axis 
               ,legend.position="bottom"                               #move legend
              ,legend.key = element_rect(fill = 'white', color="white", size = 0.0)    #fill background and remove 
              ,legend.key.size = unit(0.5,"cm")    #fill background and remove border of "key"
              ,legend.title = element_blank()                           #remove empty title entirely
              ,plot.margin = margin(t=0, r=0.35, b=0, l=0, "cm")                   #reduce plot margin ()
              ,legend.margin = margin(t=0, r=0, b=0, l=0, "cm")  
            )+
     scale_fill_manual(values=viridis_pal(option="A")(5)[c(1,3,4,2)])+
     scale_color_manual(values=viridis_pal(option="A")(5)[c(1,3,4,2)]) +  
    scale_shape_manual(values=c(21:24))

ggsave(p,filename="Figure1_2.pdf", width=8.5, height=max_column_height, units="cm")


```



#### Plot 1_1


```{r thumbnails=FALSE}

knitr::include_graphics("./Figure1_1.pdf")

```

#### Plot 1_2

```{r thumbnails=FALSE}

knitr::include_graphics("./Figure1_2.pdf")

```


# Additional resources

- https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html
- http://brooksandrew.github.io/simpleblog/articles/advanced-data-table/


#  Session Info

```{r}

sessionInfo()

```
