---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r substr((Sys.Date()),1,4)`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"    
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 009c Advanced Box Plots


## Package installation

- Copy and paste the following code in the R-console to install necessary packages.


```{r message=FALSE}

usePackage <- function(p)   
{ 
  if (!is.element(p, installed.packages()[,1])) 
    install.packages(p, dep = TRUE, repos="https://cloud.r-project.org/", lib=.libPaths()[1]) 
  require(p, character.only = TRUE) 
} 
  
usePackage("ggplot2")
usePackage("data.table")
usePackage("lvplot")   #letter value plot / boxenplot
usePackage("gghalves") #half_plots
usePackage("ggdist")   #distribution plots
usePackage("patchwork")#combine plots 
usePackage("ggridges") #ridgelines

```



## Load data and subset

```{r message=FALSE}

iris <- as.data.table(iris)


```


# Why boxplot ?

- to compare multiple groups
- **and** ideally also the data distribution


# Good practice

... but not covered here

- not-use for $<=$ 5 data points
- annotate number of observations (per plot; if these differ between variants)
  - possibility to scale by 'number'
-  


# Boxplot and data

```{r}

ggplot(data=iris, aes(x=Species,y=Sepal.Width))+
  geom_boxplot()+
  geom_jitter( height = 0,alpha=0.5)   #do not allow random positions in y-direction


```


```{r}

library("gghalves")
#https://cran.r-project.org/web/packages/gghalves/vignettes/gghalves.html

ggplot(iris, aes(x = Species, y = Sepal.Width)) +
  geom_half_boxplot() +
  geom_half_point(transformation = position_jitter(height = 0, width = 0.1, seed = 1),alpha=0.5)


ggplot(iris, aes(x = Species, y = Sepal.Width)) +
  geom_half_boxplot() +
  geom_dotplot(binaxis = "y", method="histodot", stackdir="up",binwidth=0.1,alpha=0.5)


ggplot(iris, aes(x = Species, y = Sepal.Width)) +
  geom_half_violin() +
  geom_dotplot(binaxis = "y", method="histodot", stackdir="up",binwidth=0.1,alpha=0.5)

```


# Re-order Boxplot

```{r}
# Patchwork / quick and dirty

punsort <- ggplot(iris, aes(x = Species, y = Sepal.Width)) +
  geom_boxplot()+
  ggtitle("sort by Species (default)")

preord <- ggplot(iris, aes(x = reorder(Species,Sepal.Width), y = Sepal.Width)) +
  geom_boxplot()+ 
  ggtitle("sort by mean sepal width")


punsort | preord

```


# Vertical vs. horizontal

- consider:
  - *flip*: long/multiple categories (names) on the 'x-axis' 
  - *not flip*: chronological order 


```{r}
# How to:
# with coord flip
ggplot(iris, aes(x = Species, y = Petal.Width)) +
  geom_boxplot() +
  coord_flip()

# by chaning x/y input 
ggplot(iris, aes(y = Species, x = Petal.Width)) +
  geom_boxplot() 

```





# Boxenplot / Letter value plot

- Heike Hofmann, Hadley Wickham & Karen Kafadar (2017) Letter-Value Plots: Boxplots for Large Data, *Journal of Computational and Graphical Statistics*, 26:3, 469-477, DOI: https://doi.org/10.1080/10618600.2017.1305277
- Preprint: https://vita.had.co.nz/papers/letter-value-plot.html
- https://www.r-project.org/conferences/useR-2006/Slides/HofmannEtAl.pdf
- for large data sets, as number of outliers in a boxplot scale with data size  


```{r message=FALSE}

library("lvplot")

boxenPlot0 <- 
  ggplot(iris, aes(x=Species,y=Petal.Length))+
      geom_lv()+
      ggtitle("default")

# specify number of 'boxen'

boxenPlot_k1 <- 
  ggplot(iris, aes(x=Species,y=Petal.Length))+
      geom_lv(k=1)+ #only median
      ggtitle("k=1; only median")   

boxenPlot_k2 <- 
  ggplot(iris, aes(x=Species,y=Petal.Length))+
      geom_lv(k=2,color="black")+  #median + 50% / color to see median
      ggtitle("k=2; median + 50 %")     

#Letters represent ‘median’, ‘fourths’, ‘eights’ etc.    
boxenPlot_k3_fill <- 
  ggplot(iris, aes(x=Species,y=Petal.Length))+
      geom_lv(k=3,aes(fill = ..LV.. ))  +
      ggtitle("k=3; median + 50 % + 25%")    

boxenPlot0 + boxenPlot_k1 + boxenPlot_k2 + boxenPlot_k3_fill



```

## Boxenplot with data



```{r}

ggplot(iris, aes(x=Species,y=Petal.Length))+
      geom_lv(k=5,aes(fill=Species),alpha=0.75)  +
      geom_jitter(height=0,width=0.075)

```



## Comparing distributions


```{r}
# generating new theme
theme_clean <- function(base_size = 9,
                      base_family = ""){
  theme_bw(base_size = base_size, 
                base_family = base_family) %+replace%
    theme(
      axis.title.x=element_blank(),
      axis.text.y=element_blank(),
      axis.title.y=element_blank(),
      complete = TRUE
    ) 
}

colorblindsafecolors <- c("#000000","#D55E00","#0072B2","#009E73","#CC79A7","#56B4E9","#E69F00","#F0E442")[c(6:5,1:4,7:8)]

# Simulate Data
set.seed(1)
data_normal <- data.table(x="obs",y=rnorm(n=10000,mean=0,sd=1))
data_exponential <- data.table(x="obs",y=rexp(n=10000,rate=1))
data_uniform <- data.table(x="obs",y=runif(n=10000,min=0,max=1))

# Create plots
#Boxplots
BN <- ggplot(data_normal, aes(x=x,y=y))+ geom_boxplot(fill=colorblindsafecolors[4])  + theme_clean() + coord_flip() + ggtitle("Normal(0,1)")
BE <- ggplot(data_exponential, aes(x=x,y=y))+ geom_boxplot(fill=colorblindsafecolors[5]) + theme_clean() + coord_flip()  + ggtitle("Exponential(1)")
BU <- ggplot(data_uniform, aes(x=x,y=y))+ geom_boxplot(fill=colorblindsafecolors[6])  + theme_clean() + coord_flip() + ggtitle("Uniform(0,1)")

# Violin plots
VN <- ggplot(data_normal, aes(x=x,y=y))+ geom_violin(fill=colorblindsafecolors[4],draw_quantiles=c(0.25,0.75),linetype="dashed") +
                                         geom_violin(fill="transparent",draw_quantiles=c(0.5))  + theme_clean() + coord_flip()
VE <- ggplot(data_exponential, aes(x=x,y=y))+ geom_violin(fill=colorblindsafecolors[5],draw_quantiles=c(0.25,0.75),linetype="dashed") +
                                              geom_violin(fill="transparent",draw_quantiles=c(0.5))  + theme_clean() + coord_flip()
VU <- ggplot(data_uniform, aes(x=x,y=y))+ geom_violin(fill=colorblindsafecolors[6],draw_quantiles=c(0.25,0.75),linetype="dashed") +
                                          geom_violin(fill="transparent",draw_quantiles=c(0.5))  + theme_clean() + coord_flip()
# Letter Value plots
LN <- ggplot(data_normal, aes(x=x,y=y))+ geom_lv(fill=colorblindsafecolors[4],color="black")  + theme_clean() + coord_flip()
LE <- ggplot(data_exponential, aes(x=x,y=y))+ geom_lv(fill=colorblindsafecolors[5],color="black") + theme_clean() + coord_flip() 
LU <- ggplot(data_uniform, aes(x=x,y=y))+ geom_lv(fill=colorblindsafecolors[6],color="black")  + theme_clean() + coord_flip()


# Combine Plots
(BN|BE|BU)/(VN|VE|VU)/(LN|LE|LU)


```










# Additional packages

-  `gghalves`  https://erocoar.github.io/gghalves/ 
-  `ggdist`   https://mjskay.github.io/ggdist/

```{r}

clim_lt <- fread("clim_DWD_Geisenheim_historic_daily.csv")

clim_lt[,"Month":=factor(month(MESS_DATUM))]

library("ggdist")
#TMK.Lufttemperatur

ggplot(clim_lt, aes(y=Month,x=TMK.Lufttemperatur))+ 
  stat_halfeye()

# chronological [maybe better horizontal alignment]
ggplot(clim_lt, aes(x=Month,y=TMK.Lufttemperatur))+ 
  stat_halfeye()

# Combine with 'classic' boxplot
ggplot(clim_lt, aes(x=Month,y=TMK.Lufttemperatur))+ 
  stat_interval(point_interval = median_qi,.width = c(0.5, 0.8, 0.95),alpha=0.5)+
  geom_boxplot(alpha=0.25)


# Combine some more ... | needs some fine tuning
ggplot(clim_lt, aes(x=Month,y=TMK.Lufttemperatur))+ 
  geom_half_violin(alpha=0.25,color=NA,fill="black")+
  stat_interval(point_interval = median_qi,.width = c(0.5, 0.8, 0.95),alpha=0.5)+
  #stat_summary( geom = "point",  fun = "median", color = "black", shape = 21, size=3)+
  stat_pointinterval(point_interval = median_qi,.width = c(0.5, 0.8, 0.95))+
  geom_half_point(transformation = position_jitter(height = 0, width = 0.1, seed = 1),alpha=0.02)+
  scale_color_viridis_d(option="A", begin=0.25,end=0.7)+
  theme_minimal()+
  theme(panel.grid.major.x=element_blank())


# another example
# code adapted from https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-and-how-to-create-them-with-ggplot2/
ggplot(clim_lt, aes(x = Month, y = TMK.Lufttemperatur)) + 
  ggdist::stat_halfeye(
    adjust = .5, 
    width = .6, 
    .width = 0, 
    justification = -.2, 
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .15, 
    outlier.shape = NA
  ) +
  ## add justified jitter from the {gghalves} package
  gghalves::geom_half_point(
    ## draw jitter on the left
    side = "l", 
    ## control range of jitter
    range_scale = .4, 
    ## add some transparency
    alpha = .3
  ) +
  coord_cartesian(xlim = c(1.2, NA), clip = "off")


```



## `ggridges`

- 'overlapping' density plots
- https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html

```{r}

ggplot(clim_lt, aes(x = TMK.Lufttemperatur, y = Month, fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis_c(name="T (°C)",option = "C") 

ggplot(clim_lt, aes(x = TMK.Lufttemperatur, y = Month,fill = factor(stat(quantile)))) +
  stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE,  quantiles = 4, quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles")


```




# Additional resources

- https://ggforce.data-imaginist.com/reference/geom_sina.html
- https://vita.had.co.nz/papers/letter-value-plot.html
- https://chartio.com/learn/charts/box-plot-complete-guide/
- https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-and-how-to-create-them-with-ggplot2/
- https://brshallo.github.io/r4ds_solutions/07-exploratory-data-analysis.html
- https://towardsdatascience.com/letter-value-plot-the-easy-to-understand-boxplot-for-large-datasets-12d6c1279c97
- https://github.com/easystats/see
- https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html
- https://github.com/mjskay/ggdist/blob/master/figures-source/cheat_sheet-slabinterval.pdf
- https://www.data-to-viz.com/caveat/boxplot.html

#  Session Info

```{r}

sessionInfo()

```
