---
title: "Introduction to data visualization with R and ggplot2"
author: "Dr. Dominik Schmidt"
date: "`r substr((Sys.Date()),1,4)`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc_depth: 4    
    css: "./custom.css"     
    code_folding: show
---
```{r echo=FALSE}

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
  )
```


# 010 Additional Resources




## Calculating statistics in `ggplot2`

- https://ggplot2.tidyverse.org/reference/stat_summary.html
- https://data-se.netlify.com/2018/06/10/visualizing-summary-statistics-the-tidyverse-way/
- http://rstudio-pubs-static.s3.amazonaws.com/5045_1c5e8b09671544b79f6c582bd1d9324e.html
- http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization
- https://rdrr.io/github/brooksandrew/Rsenal/man/smartAgg.html


## Two Axis (secondary X/Y axis)

- sec_axis(trans = NULL, name = waiver(), breaks = waiver(),  labels = waiver())
  - https://ggplot2.tidyverse.org/reference/sec_axis.html

## Correlations / Scatter plot matrix

- https://www.r-bloggers.com/comparing-plotly-ggplotly-plot-generation-times/
- https://www.musgraveanalytics.com/blog/2018/8/24/how-to-make-ggplot2-charts-interactive-with-plotly
- https://gganimate.com/articles/gganimate.html
- https://ggvis.rstudio.com/


## Heatmaps/ Dendrograms

- https://mode.com/blog/r-ggplot-extension-packages
- https://jcoliver.github.io/learn-r/006-heatmaps.html
- http://www.opiniomics.org/you-probably-dont-understand-heatmaps/
- https://earlglynn.github.io/RNotes/package/gplots/heatmap2.html
- https://genomicsclass.github.io/book/pages/clustering_and_heatmaps.html
- http://girke.bioinformatics.ucr.edu/GEN242/pages/mydoc/Rclustering.html
- https://cran.r-project.org/web/packages/ggdendro/vignettes/ggdendro.html
- http://www.sthda.com/english/wiki/beautiful-dendrogram-visualizations-in-r-5-must-known-methods-unsupervised-machine-learning

## More add-ons/extensions/features

- https://exts.ggplot2.tidyverse.org/gallery/
- https://ggvis.rstudio.com/
- http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/
- https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html
- https://github.com/thomasp85/ggforce
- https://github.com/AtherEnergy/ggTimeSeries 
- https://pkg.earo.me/sugrrants/ 
- https://pkg.earo.me/sugrrants/articles/frame-calendar.html
- https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
- http://davidgohel.github.io/ggiraph/index.html
- http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization

## Misc

- `ggpubr` - Shortcuts for pre-designed plots -- https://rpkgs.datanovia.com/ggpubr/


## Dos and Don'ts

- http://genomicsclass.github.io/book/pages/plots_to_avoid.html
- Christa Kelleher, Thorsten Wagener, Ten guidelines for effective data visualization in scientific publications, Environmental Modelling & Software, Volume 26, Issue 6, 2011, Pages 822-827, ISSN 1364-8152, https://doi.org/10.1016/j.envsoft.2010.12.006 (http://www.sciencedirect.com/science/article/pii/S1364815210003270)
- Rougier NP, Droettboom M, Bourne PE (2014) Ten Simple Rules for Better Figures. PLoS Comput Biol 10(9): e1003833. https://doi.org/10.1371/journal.pcbi.1003833
- http://blogs.nature.com/methagora/2013/07/data-visualization-points-of-view.html?WT.mc_id=TWT_NatureMethods
- https://stat545.com/effective-graphs.html
- https://visguides.org/
- https://eagereyes.org/talk/how-do-we-know-that

## More Dimension (3D, 4D, ...)

- https://towardsdatascience.com/a-comprehensive-guide-to-the-grammar-of-graphics-for-effective-visualization-of-multi-dimensional-1f92b4ed4149
- 'Real' 3D Plots: `rgl` - interactive (https://cran.r-project.org/web/packages/rgl/vignettes/rgl.html)
- 'Real' 3D Plots: `scatterplot3d` static (https://cran.r-project.org/web/packages/scatterplot3d/vignettes/s3d.pdf)
- https://rdrr.io/rforge/rgl/man/persp3d.html
- http://www.sthda.com/english/wiki/impressive-package-for-3d-and-4d-graph-r-software-and-data-visualization
- https://ggplot2.tidyverse.org/reference/geom_tile.html


### Plot Pseudo-3d -- Colored points

```{r}
library("ggplot2")
library("data.table")
#### Create data 
# Measurements over a rectangular surface

x <- 0:10
y <- x

set.seed(2019)
value <- rnorm(mean=100,sd=10, n=length(x) * length(y))

d <- as.data.table(expand.grid(x=x,y=y))
d[,"value":=value*x/(y+1)]


```


```{r}

# Option1: Colour data points by value | viridis colors
ggplot(d,aes(x=x,y=y))+
  geom_point(aes(color=value))+
  coord_fixed()+                       #Fix aspect ratio
  scale_colour_viridis_c()


```

### Plot Pseudo-3d -- Colored raster


```{r}



ggplot(d,aes(x=x,y=y))+
  geom_raster(aes(fill=value))+        #Coordinates are in the center of the squares
  coord_fixed()+                       #Fix aspect ratio
  scale_fill_viridis_c()


ggplot(d[-13,],aes(x=x,y=y))+          #Missing data = Missing
  geom_raster(aes(fill=value))+
  coord_fixed()+                       
  scale_fill_viridis_c()


ggplot(d[-13,],aes(x=x,y=y))+
  geom_raster(aes(fill=value),interpolate=TRUE)+  #NOT very good ?! missing = zero
                                                  #AND to strong boarders
  coord_fixed()+                      
  scale_fill_viridis_c()

# google: Missing values problem [data imputation]  

```

### Real 3D


```{r}
library("rgl")   #interative 3D plots
  options(rgl.useNULL = TRUE) #only for markdown output

# Pointcloud
plot3d(d)
rgl_scene <- scene3d()        #only for markdown output
rglwidget(x=rgl_scene)        #only for markdown output

```



### Four Dimensions with a few lines of ggplot2 

- x (wt)
- y (mpg)
- color (gear)
- +size (cyl)


```{r}

ggplot(mtcars, aes(x=wt, y=mpg, color=factor(gear), size=cyl))+
        geom_point() +
        theme_minimal()


```



#  Session Info

```{r}

sessionInfo()

```
